\documentclass[proposal,english,palatino]{AIGpaper}

%%%% Package Imports %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{graphicx}					    % enhanced support for graphics
\usepackage{tabularx}				      	% more flexible tabular
\usepackage{amsfonts}					    % math fonts
\usepackage{amssymb}					    % math symbols
\usepackage{amsmath}					    % overall enhancements to math environment
\usepackage{acronym}                        % for acronym management
\usepackage{xcolor}                         % for color support

%%%% optional packages
\usepackage{tikz}                           % creating graphs and other structures
\usetikzlibrary{arrows,positioning}
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for argument
    args/.style={circle, minimum size=0.9cm,draw=black, thick,fill=white},
}

%%%% Author and Title Information %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\InputIfFileExists{../meta.tex}{}{\newcommand{\authorname}{Erika Mustermann}}
\author{\authorname}

\title{Inconsistency Measurement in Classical Planning Problems}

\begin{document}

% Acronym definitions
\acrodef{ASP}{Answer Set Programming}
\acrodef{STRIPS}{Stanford Research Institute Problem Solver}
\acrodef{PDDL}{Planning Domain Definition Language}
\acrodef{SAT}{Boolean Satisfiability}

\maketitle % prints title and author information, as well as the abstract 

% ===================== Beginning of the actual text section =====================

\section{Introduction and Motivation}

In practical applications of automated planning, systems regularly encounter unsolvable problems. A planning instance is unsolvable if no sequence of available actions can transform the initial state into a goal state \cite{Russell2016}. Since determining solvability is PSPACE-complete \cite{Bylander1994}, unsolvable instances arise naturally in complex domains rather than being merely pathological cases. Consider a logistics scenario where a delivery company needs to route packages through a network of distribution centers. If the problem specifies that a package must be delivered to New York by noon while simultaneously requiring it to be processed at a facility in Los Angeles at the same time, the planning system will simply report "unsolvable" without explaining the conflicting requirements \cite{Goebelbecker2010,Sreedharan2019}. Similarly, in robotics applications, a robot might be asked to manipulate objects in ways that violate physical constraints, or in manufacturing, production schedules might contain contradictory resource requirements \cite{Goebelbecker2010}.

Beyond recognizing unsolvability, users require explanations of why the planner failed to find a solution. Current planning systems excel at finding solutions when they exist, but provide minimal diagnostic information when problems are unsolvable \cite{Eriksson2018,Sreedharan2019}. This limitation becomes particularly problematic in complex domains where problems are automatically generated or when domain experts need to understand why their specifications cannot be satisfied \cite{Sreedharan2019}. Manual debugging of such failures is impractical for large-scale problems with hundreds of constraints and actions, yet current approaches offer no systematic alternatives \cite{Goebelbecker2010,Sreedharan2019}. The lack of quantitative diagnostic capabilities forces users into costly trial-and-error processes, significantly hindering the practical deployment of automated planning in real-world applications \cite{Eriksson2018}.

The field of inconsistency measurement in knowledge representation and reasoning offers a promising approach to address this limitation \cite{Grant2017,Thimm2016}. Rather than treating inconsistency as a binary property, these techniques quantify the degree of inconsistency within logical knowledge bases. Various measures have been developed to identify minimal inconsistent subsets, calculate distances to consistency, and count conflicting elements \cite{Grant2011,Ulbricht2016}. However, these approaches have been primarily applied to static logical knowledge bases and have not been systematically adapted to the temporal and procedural nature of planning problems \cite{Ulbricht2016,Corea2022,Condotta2024}.

While techniques for quantifying inconsistencies have been extensively developed in knowledge representation and reasoning, their systematic adaptation to planning domains remains largely unexplored. Current planning research focuses primarily on binary diagnostics (solvable vs.\ unsolvable) without providing quantitative insights into the severity, structure, or complexity of conflicts \cite{Eriksson2018}. Planning problems present unique challenges for inconsistency measurement since, unlike propositional knowledge bases, planning involves sequences of actions that transform states over time \cite{Russell2016}. Conflicts can arise not only from contradictory static constraints but also from action sequences that lead to impossible state combinations or from goals that cannot be reached given the available actions. The temporal structure means that inconsistencies might only emerge when considering specific action orderings or at particular time points in the planning horizon \cite{Condotta2024}.

This research aims to bridge inconsistency measurement and planning by exploring how techniques from knowledge representation and reasoning can be adapted to quantify the degree of inconsistency in unsolvable planning instances. In this thesis, we develop quantitative diagnostics that characterize which goals or constraints drive a task's unsolvability. The approach focuses on \ac{ASP}-based algorithms, which are particularly well-suited for this task due to their declarative modeling capabilities, built-in support for temporal and combinatorial constraints, and recent performance improvements in solving optimization problems common in planning scenarios \cite{Dimopoulos1997,DIMOPOULOS2019,Ulbricht2016,Kuhlmann2025}. If successful, such quantitative measures could provide systematic approaches to analyzing and fixing planning problems, filling a critical gap in current planning methodology.

\section{Technical Foundations and Related Work}

The following sections establish the theoretical foundation for this research by examining inconsistency measurement techniques, classical planning formalisms, and prior work connecting planning with inconsistency analysis.

\subsection{Inconsistency Measurement in Knowledge Representation and Reasoning}

Inconsistency measurement provides formal methods for quantifying the degree of inconsistency within logical knowledge bases. Rather than treating inconsistency as a binary property, these measures assign numerical values that reflect the severity or extent of conflicts, enabling comparative analysis and systematic approaches to conflict resolution.

Among the various approaches that have been developed, three main categories are particularly interesting for planning applications:

% Todo: correct descriptions of measures
% Die Beschreibung der Inkonsistenzmaße ist teilweise nicht ganz korrekt und auch nicht so ideal strukturiert.
%     Das Hitting Set-Maß steht relativ für sich alleine; es gibt nicht direkt ähnliche Maße.
%     "Contension" als Benennung zu wählen, ist auch nicht so gut; es gibt ja nur das eine Contension-Maß. Es gibt allerdings noch andere, die auch auf paraconsistent semantics basieren.
%     Beim Hit-Distance-Maß stellt sich die Frage, warum es nicht den Distanz-basierten Maßen zugeordnet wird?
%     Das Hitting Set-Maß ist nicht ganz korrekt beschrieben: "that "hit" every conflict" -> Ein Hitting Set ist in diesem Kontext eine Menge von Interpretationen, sodass jede Formel in der gegebenen Knowledge Base von mind. einer dieser Formeln erfüllt wird. 
%     "Contension-based measures quantify inconsistency by finding the smallest set of propositions that must be marked "both true and false" to model the data, then counting those conflicting statements" 
%         Was bedeutet denn "to model the data"? Sollte eigentlich eher heißen "to render the knowledge base consistent under Priest's three-valued logic" oder "under paraconsistent semantics"
%         und "counting those conflicting statements" ist so gesehen auch nicht ganz richtig (oder zumindest ungenau) -- es wird gezählt, wie viele Atome in der Signatur auf both gesetzt werden müssen"
{\color{red}\textbf{Distance-based measures} quantify inconsistency by calculating how far a knowledge base is from being consistent. They identify the minimal modifications needed to resolve all conflicts, measuring this "distance to consistency" in different ways. For example, the Max-Distance measure finds the smallest number of formula modifications needed if one can change any single formula as much as necessary, while Sum-Distance counts the total number of individual changes across all formulas \cite{Grant2017}. These measures provide an intuitive notion of repair cost, directly answering "how much would we need to change to fix this?"

\textbf{Hitting-set-based measures} quantify inconsistency by identifying all minimal conflicting subsets of a knowledge base, then calculating the smallest sets of formulas (hitting sets) that "hit" every conflict. These measures use the size or overlap of those hitting sets to estimate how deeply entangled the inconsistencies are. The Hitting-Set and Hit-Distance measures fall into this category and are considered computationally reasonable to calculate \cite{Thimm2016}. These measures focus on the structural aspects of inconsistency by analyzing which formulas participate in conflicts and how conflicts interconnect.

\textbf{Contension-based measures} quantify inconsistency by finding the smallest set of propositions that must be marked "both true and false" to model the data, then counting those conflicting statements. This gives a clear, diagnostic indication of where and how many contradictions occur. The Contension measure is one of the most studied in the literature \cite{Grant2011}. These measures provide a direct count of problematic elements, making them particularly interpretable for diagnostic purposes.}

Recent advances show that \ac{ASP}-based algorithms for computing inconsistency measures can outperform \ac{SAT}-based approaches for many measures \cite{Kuhlmann2025}. This suggests significant potential for planning applications, as \ac{ASP} offers several key advantages: its declarative nature allows natural modeling of planning constraints and temporal relationships, and its support for non-monotonic reasoning aligns well with planning's inherent non-monotonic aspects such as the frame problem and default assumptions. Crucially, \ac{ASP} can naturally represent temporal sequences through additional time arguments in atoms (e.g., \texttt{holds(fluent, T)}) and sequential rules, making it particularly well-suited for modeling planning scenarios with their inherent temporal structure and action sequences.

This temporal modeling approach was established by \cite{Dimopoulos1997}, who first demonstrated how to encode \ac{STRIPS}-like planning problems as non-monotonic logic programs with time-indexed actions and fluents, showing that answer sets correspond directly to valid plans. Building on this foundation, the approach has evolved significantly, with modern tools such as plasp 3 \cite{DIMOPOULOS2019} now providing efficient translations from standard \ac{PDDL} planning instances to \ac{ASP} encodings. These contemporary systems maintain the core temporal modeling principles while extending support to handle complex planning features through domain-independent \ac{ASP} rule sets.

Additionally, Ulbricht et al. \cite{Ulbricht2016} developed inconsistency measures specifically for \ac{ASP}, demonstrating that the monotonicity postulate does not hold in non-monotonic logics. This insight may prove useful when adapting measures to planning, where reasoning often involves non-monotonic aspects such as the frame problem and default assumptions about action effects.

\subsection{Classical Planning}

Classical planning operates on a state-based formalism where the world is represented as a set of propositions that can be true or false in any given state. Actions are defined by their preconditions and effects: what must be true for the action to be applicable, and how the action changes the world state. The \ac{STRIPS} representation is foundational for many modern planning systems \cite{Russell2016}, providing a simple yet powerful framework for modeling planning domains.

A planning problem consists of three components: an initial state describing the starting configuration, a set of available actions, and a goal specification defining the desired outcome. The planning task is to find a sequence of actions that, when executed from the initial state, results in a state satisfying the goal conditions.

Planning problems become unsolvable when no sequence of applicable actions can transform the initial state into a goal state. This can occur for several reasons: goals that are mutually exclusive and require contradictory states, required resources that are unavailable or insufficient, action preconditions that can never be satisfied, or action sequences that lead to dead-end states from which the goal cannot be reached. Determining whether a plan exists is computationally expensive, which has motivated extensive research into efficient algorithms and search techniques.

\ac{PDDL} serves as the standard language for describing planning problems and is used in international planning competitions \cite{Ghallab1998}. \ac{PDDL} provides a structured way to specify domains and problems, making it essential for practical applications. However, given the complexity of \ac{PDDL}'s various extensions and features, this thesis initially focuses on \ac{STRIPS}-based formulations due to their foundational simplicity and well-understood semantics, which allow for systematic adaptation of inconsistency measures without the complications of advanced \ac{PDDL} constructs.

Modern planning systems employ various search strategies such as forward state-space search, backward regression, and plan-space search, along with sophisticated guidance techniques. However, when these systems encounter an unsolvable problem, they typically only report failure without providing analysis of why the problem is unsolvable, which conflicts might be responsible, or what modifications could potentially make the problem solvable. This diagnostic gap represents a significant limitation for users trying to understand and fix problematic planning instances. % Todo: source

\subsection{Planning and Inconsistency: Foundation Work}

The connection between planning and inconsistency has been explored from several perspectives, though quantitative inconsistency measurement remains largely unexplored in the planning literature.

Eriksson et al. \cite{Eriksson2018} established important theoretical groundwork by connecting unsolvable planning problems to logical inconsistency. Their work proposes a proof system for demonstrating unsolvability in planning, examining why certain planning tasks cannot be solved using STRIPS-based approaches. While this research formalizes how planning failures relate to logical inconsistency, it focuses on proving unsolvability rather than measuring degrees of inconsistency.

From a diagnostic perspective, Göbelbecker et al. \cite{Goebelbecker2010} introduced the concept of "excuses" for unsolvable planning tasks. An excuse represents a minimal modification required to make an unsolvable problem solvable, such as removing or altering a specific goal or constraint. This approach is conceptually similar to identifying minimal inconsistent subsets in logical knowledge bases, which is the type of structural analysis that underlies hitting-set-based measures. However, excuses provide only binary guidance about what to remove rather than quantitative measures of inconsistency severity, and lack the comparative analysis capabilities that would allow users to understand the relative complexity of different conflicts or prioritize repair strategies.

More recently, Sreedharan et al. \cite{Sreedharan2019} developed methods for producing explanations that help users understand why planning tasks are unsolvable, focusing on the identification of unreachable subgoals. Their approach emphasizes explanatory aspects rather than quantitative measurement, but demonstrates the practical value of systematic analysis when planning problems are unsolvable. While valuable for understanding individual cases of unsolvability, these explanation methods do not provide the quantitative measures needed for systematic problem analysis, automated repair, or comparative evaluation of different unsolvable instances.

These research efforts highlight both the need for better diagnostic tools and the limitations of current approaches. Unlike static knowledge bases, planning problems involve temporal sequences and procedural constraints that create new types of conflicts. While existing work provides valuable insights into individual aspects of unsolvable planning problems, there is a clear gap in quantitative approaches that can systematically measure and compare the degree of inconsistency across different planning instances. This thesis addresses this gap by developing the first comprehensive framework for quantitative inconsistency measurement in planning, building upon but extending beyond the binary diagnostic approaches of prior work.

\section{Problem Statement}

When current planning systems encounter unsolvable problems, they provide very limited diagnostic information. This creates several challenges for users and researchers. Users have little insight into why a given planning problem is unsolvable, and there is no systematic way to analyze the structure of an unsolvable planning instance. This makes it difficult to identify what minimal changes could make the problem solvable, as different types of unsolvable planning problems are not distinguished or quantified in any way. The lack of quantitative diagnostic capabilities forces users into trial-and-error processes, significantly limiting the practical applicability of automated planning in complex real-world scenarios.

\section{Research Questions and Objectives}

The central research question of this thesis is:

\begin{quote}
    How can inconsistency measures from knowledge representation and reasoning be meaningfully adapted to classical planning to provide diagnostic value for unsolvable planning problems?
\end{quote}

This central question is addressed through the following research questions and their corresponding objectives:

\subsection{Research Question 1: Theoretical Adaptation}

\begin{quote}
    Which inconsistency measures from knowledge representation and reasoning can be adapted to classical planning, and what postulates do they retain?
\end{quote}

\begin{enumerate}
    \item[(1a)] Survey and categorize existing inconsistency measures from knowledge representation and reasoning.
    \item[(1b)] Analyze the structural differences between knowledge bases and planning problems, identifying adaptation challenges.
    \item[(1c)] Develop formal adaptations of 2-3 selected inconsistency measures to \ac{STRIPS}-based planning formulations.
    \item[(1d)] Examine which postulates are preserved or modified in the planning context.
    \item[(1e)] Provide formal definitions and examples of the adapted measures.
\end{enumerate}

\subsection{Research Question 2: Computational Implementation}

\begin{quote}
    How can the adapted measures be efficiently computed for planning problems using Answer Set Programming?
\end{quote}

\begin{enumerate}
    \item[(2a)] Design \ac{ASP}-based encodings for the adapted inconsistency measures.
    \item[(2b)] Develop algorithms that handle standard planning problem formats, including \ac{STRIPS} and optionally basic \ac{PDDL} representations.
    \item[(2c)] Implement a prototype system that can parse planning problems and compute the adapted measures.
    \item[(2d)] Analyze computational complexity and if possible optimize \ac{ASP} encodings for scalability in planning domains.
\end{enumerate}

\subsection{Research Question 3: Practical Evaluation (Optional)}

\begin{quote}
    Do the adapted measures provide practical diagnostic value for understanding and analyzing unsolvable planning problems?
\end{quote}

\begin{enumerate}
    \item[(3a)] Establish an experimental framework using benchmark planning problems from established collections.
    \item[(3b)] Evaluate computational feasibility and scalability of the implemented measures on problems of varying complexity.
    \item[(3c)] Analyze the diagnostic utility of the measures for distinguishing different types of unsolvable planning problems.
    \item[(3d)] Assess whether the measures provide actionable insights for problem repair and modification strategies.
\end{enumerate}

\textbf{Note:} Research Question 3 is marked as optional due to the substantial scope of Questions 1 and 2. The theoretical development and implementation phases constitute the core contribution of this work, and the evaluation component will be pursued if time and complexity constraints permit.

\section{Planned Structure}

The thesis will be organized into seven chapters:

\begin{description}
    \item[Chapter 1: Introduction] Overview of the problem domain, motivation, and summary of contributions.
    \item[Chapter 2: Foundations] Background on classical planning formalisms, inconsistency measurement theory, and \ac{ASP}.
    \item[Chapter 3: Related Work] Survey of existing work on planning diagnostics and inconsistency measures.
    \item[Chapter 4: Theoretical Adaptation] Formal adaptation of selected inconsistency measures to planning, with definitions and examples.
    \item[Chapter 5: Implementation] \ac{ASP}-based algorithms and prototype development.
    \item[Chapter 6: Evaluation] Experimental setup, benchmarking results, and analysis.
    \item[Chapter 7: Conclusion] Summary of findings, limitations, and future work.
\end{description}

\section{Timeline}

The project is planned over approximately six months, with overlapping phases:

\begin{description}
    \item[Months 1-2:] Literature review and theoretical foundations (Chapters 2-3).
    \item[Months 2-4:] Theoretical development and formal adaptation of measures (Chapter 4).
    \item[Months 3-5:] Implementation of \ac{ASP}-based algorithms and prototype development (Chapter 5).
    \item[Months 4-6:] Experimental evaluation and thesis write-up (Chapters 1, 5-6).
\end{description}

\newpage

% References
\addreferences

\end{document}
