\documentclass[master,twoside,binding,palatino,english]{AIGpaper}
% Please read the README.md file for additional information on the parameters and overall usage of AIGpaper

\usepackage{graphicx}					    % enhanced support for graphics
\usepackage{tabularx}				      	% more flexible tabular
\usepackage{amsfonts}					    % math fonts
\usepackage{amssymb}					    % math symbols
\usepackage{amsmath}					    % overall enhancements to math environment
\usepackage{amsthm}                         % theorem environments
\usepackage{acronym}                        % for acronym management
\usepackage{url}                            % URLs in bibliography
\usepackage{booktabs}                       % professional table rules
\usepackage{float}                          % [H] float specifier

\theoremstyle{definition}
\newtheorem{defn}{Definition}
\newtheorem{example}{Example}

\theoremstyle{plain}
\newtheorem{prop}{Proposition}
\newtheorem{thm}{Theorem}
\newtheorem{cor}{Corollary}

% optional packages
\usepackage{tikz}                           % creating graphs and other structures
\usetikzlibrary{arrows,positioning,fit,backgrounds}
\usepackage{forest}                         % tree diagrams (taxonomy figure)
\usepackage{listings}                       % code listings (ASP encoding)
\usepackage{needspace}
\usepackage{pgfplots}
\usepackage[hidelinks]{hyperref}
\lstset{
    basicstyle=\small\ttfamily,
    keywordstyle=\bfseries,
    commentstyle=\itshape\color{gray},
    frame=single,
    breaklines=true,
    captionpos=b,
    numbers=left,
    numberstyle=\tiny\color{gray},
    xleftmargin=2em,
    framexleftmargin=1.5em,
}
\BeforeBeginEnvironment{lstlisting}{\Needspace{5\baselineskip}}
\lstdefinestyle{asp}{
    language={},
    alsoletter={\#},
    morekeywords={not, true, false, \#show, \#const, \#minimize, \#maximize},
    keywordstyle=\bfseries,
    morecomment=[l]{\%},
}
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for argument
    args/.style={circle, minimum size=0.9cm,draw=black, thick,fill=white},
}
\pgfplotsset{compat=1.18}

% prevent widows and orphans (also before/after display math)
\widowpenalty=10000
\clubpenalty=10000
\displaywidowpenalty=10000
\predisplaypenalty=10000

% allow line breaks after escaped underscores in text (e.g. has\_key)
\let\origunderscore\_
\renewcommand{\_}{\origunderscore\discretionary{}{}{}}

% === set information ===
% if meta.tex does not exist, use default values
\InputIfFileExists{../meta.tex}{}{
    \newcommand{\authorname}{Erika Mustermann}
    \newcommand{\myDegreecourse}{Informatik}
    \newcommand{\reviewername}{Prof.\ Dr.\ Matthias Thimm}
    \newcommand{\advisorname}{Max Mustermann}
    \newcommand{\myCity}{}
}
\author{\authorname}

\title{Inconsistency Measurement for Classical Planning Problems}

\degreecourse{\myDegreecourse}

\firstreviewer{\reviewername}
\firstreviewerinfo{Artificial Intelligence Group}

\advisor{\advisorname}
\advisorinfo{Artificial Intelligence Group}

\germanabstract{
Klassische Planungssysteme melden lediglich, falls ein Problem unlösbar ist, und liefern keine weiteren Erklärungen. Diese Arbeit überträgt Inkonsistenzmessung aus der Aussagenlogik auf die klassische Planung und stellt strukturelle Diagnostiken der Unlösbarkeit bereit. Eine aussagenlogische Kodierung mit dem Contension-Maß liefert uninformative Ergebnisse und motiviert Maße, die direkt auf Zielen, Operatoren und erreichbaren Zuständen operieren. Drei planungsnative Diagnoseprofile mit insgesamt sechs Maßen werden vorgeschlagen. Das Unerreichbarkeits\-profil zählt Zielpropositionen, die in keinem erreichbaren Zustand vorkommen. Das Zielmutex-Profil erkennt operatorinduzierte Ausschlüsse, bei denen Ziele einzeln erreichbar, aber nie gemeinsam erfüllbar sind. Das Zielsequenzierungskonflikt-Profil identifiziert irreversible Ausschlüsse, bei denen das Erreichen eines Ziels ein anderes dauerhaft blockiert. Jedes Profil ist durch einen Planlängenhorizont parametrisiert, wird gegen planungsadaptierte Rationalitätspostulate analysiert und ist PSPACE-vollständig. Die Profile werden mittels Answer Set Programming unter Verwendung von \textit{plasp} und Clingo-Brave-Reasoning implementiert. Eine experimentelle Auswertung auf Benchmarks der Unsolvability Competition der International Planning Competition 2016 sowie der Eriksson-Sammlung zeigt, dass die Profile strukturell unterschiedliche Kategorien unlösbarer Instanzen unterscheiden, und offenbart ein Spannungsverhältnis zwischen Genauigkeit und Abdeckung, das der horizontbeschränkten Analyse inhärent ist.
}
\englishabstract{
Classical planners report only when a problem is unsolvable, leaving the cause unexplained. This thesis adapts inconsistency measurement from propositional logic to classical planning, providing structural diagnostics of unsolvability. Encoding into propositional knowledge bases with the contension measure yields uninformative results, motivating measures that operate directly on goals, operators, and reachable states. Three planning-native diagnostic profiles comprising six measures in total are proposed. The Unreachability Profile counts goal propositions absent from all reachable states. The Goal Mutex Profile detects operator-induced mutual exclusions where goals are individually achievable but never satisfiable in a common reachable state. The Goal Sequencing Conflict Profile identifies irreversible exclusions where achieving one goal permanently blocks another. Each profile is parameterized by a plan-length horizon, analyzed against planning-adapted rationality postulates, and shown to be PSPACE-complete. The profiles are implemented in Answer Set Programming using \textit{plasp} and Clingo brave reasoning. An experimental evaluation on the 2016 International Planning Competition's Unsolvability Competition and the Eriksson benchmark collection shows that the profiles distinguish structural categories of unsolvable instances and reveal a soundness-coverage tradeoff inherent to horizon-bounded analysis.
}

\begin{document}

% Acronym definitions
\acrodef{API}{application programming interface}
\acrodef{ASP}{Answer Set Programming}
\acrodef{CLI}{command-line interface}
\acrodef{IPC}{International Planning Competition}
\acrodef{PDDL}{Planning Domain Definition Language}
\acrodef{SAT}{Boolean Satisfiability}
\acrodef{STRIPS}{Stanford Research Institute Problem Solver}

\pagenumbering{roman}
\maketitle % prints the cover page an empty page if two-sided print

\tableofcontents
\cleardoublepage

\listoffigures
\listoftables
\cleardoublepage

\pagenumbering{arabic}

% ========== beginning of the actual text section ===========
\input{masterthesis_sections/introduction.tex}
\input{masterthesis_sections/foundations.tex}
\input{masterthesis_sections/related_work.tex}
\input{masterthesis_sections/theoretical_adaptation.tex}
\input{masterthesis_sections/implementation.tex}
\input{masterthesis_sections/evaluation.tex}
\input{masterthesis_sections/conclusion.tex}

% References
\cleardoublepage
\addreferences

% Appendices
\cleardoublepage
\input{masterthesis_sections/appendix.tex}

\end{document}
