\section{Foundations}

This section establishes the theoretical foundations for adapting paraconsistent inconsistency measures to classical planning problems. The chapter begins with formal definitions for knowledge bases and inconsistency measurement, with particular emphasis on paraconsistent approaches. This is followed by an examination of classical planning formalisms and the specific challenges that arise when adapting inconsistency measures to the planning domain.

\subsection{Mathematical Foundations of Inconsistency Measurement}

Inconsistency measurement provides formal methods for quantifying the degree of inconsistency within logical knowledge bases. Rather than treating inconsistency as a binary property, these measures assign numerical values that reflect the severity or extent of conflicts, enabling comparative analysis and systematic approaches to conflict resolution~\cite{Thimm2019}.

\subsubsection{Formal Definition of Knowledge Bases}

The fundamental concept of a knowledge base in propositional logic is defined as follows.

\begin{defn}
    Let $\mathcal{L}$ be a propositional language over a finite alphabet $\Sigma$ of propositional atoms. A \emph{knowledge base} $K$ is a finite set of propositional formulas from $\mathcal{L}$, i.e., $K = \{\phi_1, \phi_2, \ldots, \phi_n\}$ where each $\phi_i \in \mathcal{L}$.
\end{defn}

A knowledge base $K$ is \emph{consistent} if there exists an interpretation $\mathcal{I}$ such that $\mathcal{I} \models \phi$ for every $\phi \in K$. Otherwise, $K$ is \emph{inconsistent}. The set of all interpretations of $\mathcal{L}$ is denoted as $\text{Mod}(\mathcal{L})$. The set of models of a formula set $K$ is defined as:
\begin{equation*}
    \text{Mod}(K) = \{\mathcal{I} \in \text{Mod}(\mathcal{L}) : \mathcal{I} \models \phi \text{ for all } \phi \in K\}
\end{equation*}

\subsubsection{Formal Definition of Inconsistency Measures}

An inconsistency measure provides a systematic way to assign numerical values reflecting the degree of inconsistency.

\begin{defn}
    An \emph{inconsistency measure} is a function $I : \mathcal{K} \to \mathbb{R}_{\geq 0}$ that maps every knowledge base from a given class $\mathcal{K}$ to a non-negative real number, where $I(K) = 0$ if and only if $K$ is consistent.
\end{defn}

Inconsistency measures are typically required to satisfy certain rationality postulates that capture intuitive properties of inconsistency quantification~\cite{Hunter2008,Thimm2019}. However, these postulates may vary depending on the specific context or domain of application:

\begin{enumerate}
    \item \textbf{Consistency:} $I(K) = 0$ if and only if $K$ is consistent.
    \item \textbf{Monotonicity:} If $K \subseteq K'$, then $I(K) \leq I(K')$.
    \item \textbf{Dominance:} If $K \subseteq K'$ and $K'$ is inconsistent, then $I(K) < I(K')$.
    \item \textbf{Free Formula Independence:} $I(K \cup \{\phi\}) = I(K)$ if $\phi$ is not involved in any inconsistency of $K \cup \{\phi\}$.
\end{enumerate}

\subsubsection{Approaches to Inconsistency Measurement}

Several approaches to inconsistency quantification have been developed in the literature, each capturing different aspects of logical conflicts~\cite{Thimm2019}. Syntactic approaches focus on structural properties such as minimal inconsistent subsets and maximal consistent subsets, typically counting or analyzing these sets to quantify inconsistency. Other approaches include distance-based measures that compute the minimal modifications needed to restore consistency, and hitting-set measures that seek minimal sets of interpretations satisfying all formulas~\cite{Grant2013,Grant2017,Thimm2016}.

\subsubsection{Paraconsistent Semantics and the Contension Measure}

This thesis focuses on paraconsistent approaches to inconsistency measurement, which offer a fundamentally different perspective by examining conflicts at the propositional level rather than the formula level. The contension-based measure employs paraconsistent semantics, specifically Priest's logic of paradox~\cite{Priest1979}, which introduces a third truth value to handle contradictions.

A three-valued interpretation $\upsilon : \text{At} \to \{T, F, B\}$ extends classical interpretations by adding the truth value $B$ (``both'') to represent propositions that are simultaneously true and false. The function $\upsilon$ extends to arbitrary formulas according to the truth tables in Table~\ref{tab:three-valued-logic}. A key property is that negation preserves the conflicting value: $\upsilon(\neg \phi) = B$ when $\upsilon(\phi) = B$, reflecting that if a proposition is both true and false, its negation is also both true and false.

\begin{table}[ht]
    \centering
    \begin{tabular}{cc|cc|c|c}
        \hline
        $\phi$ & $\psi$ & $\phi \land \psi$ & $\phi \lor \psi$ & $\phi$ & $\neg \phi$ \\
        \hline
        T      & T      & T                 & T                & T      & F           \\
        T      & B      & B                 & T                & B      & B           \\
        T      & F      & F                 & T                & F      & T           \\
        B      & T      & B                 & T                &        &             \\
        B      & B      & B                 & B                &        &             \\
        B      & F      & F                 & B                &        &             \\
        F      & T      & F                 & T                &        &             \\
        F      & B      & F                 & B                &        &             \\
        F      & F      & F                 & F                &        &             \\
        \hline
    \end{tabular}
    \caption{Truth tables for three-valued logic}
    \label{tab:three-valued-logic}
\end{table}

An interpretation $\upsilon$ satisfies a formula $\phi$ (written $\upsilon \models_3 \phi$) if $\upsilon(\phi) = T$ or $\upsilon(\phi) = B$. This paraconsistent satisfaction relation allows formulas to be satisfied even when they evaluate to the conflicting value $B$. Under this semantics, every knowledge base possesses at least one satisfying interpretation: assigning $B$ to all propositions trivially satisfies any $K$, even when $K$ is classically inconsistent.

To illustrate the computation of minimal $B$-assignments, consider the knowledge base
\begin{equation*}
    K = \{\text{sunny}, \neg\text{sunny}, \text{hot}\}\,.
\end{equation*}
The goal is to find an interpretation $\upsilon$ minimizing the number of propositions assigned $B$.

Attempting a classical interpretation with $\upsilon(\text{sunny}) = T$ and $\upsilon(\text{hot}) = T$ yields $\upsilon(\neg\text{sunny}) = F$. This violates the requirement $\upsilon \models_3 \neg\text{sunny}$, so this interpretation does not satisfy $K$.

However, the interpretation with $\upsilon(\text{sunny}) = B$ and $\upsilon(\text{hot}) = T$ successfully satisfies all formulas:
\begin{itemize}
    \item $\upsilon(\text{sunny}) = B$ satisfies \textit{sunny}
    \item $\upsilon(\neg\text{sunny}) = B$ satisfies $\neg$\textit{sunny}
    \item $\upsilon(\text{hot}) = T$ satisfies \textit{hot}
\end{itemize}
This assignment uses exactly one $B$-valued proposition and cannot be reduced further.

The contension measure~\cite{Grant2011} formalizes this minimization process:
\begin{equation*}
    I_c(K) = \min\{|\upsilon^{-1}(B) \cap \text{At}| \mid \upsilon \models_3 K\}
\end{equation*}

This measure differs fundamentally from structural approaches by examining conflicts at the propositional level rather than the formula level. This distinction is particularly relevant when conflicts involve formulas of different granularities. For instance, while $\{a, \neg a\}$ and $\{a \land b \land c, \neg a \land \neg b \land \neg c\}$ both contain a single minimal inconsistent subset, the contension measure yields $I_c = 1$ and $I_c = 3$ respectively, reflecting that the latter involves conflicts across three propositions rather than one~\cite{Thimm2019}. This proposition-centric view provides finer-grained distinctions in inconsistency severity.

\subsubsection{Domain Adaptations}

While originally developed for propositional knowledge bases, inconsistency measurement techniques have been successfully adapted to diverse logical domains, establishing important precedents for domain-specific adaptations. These adaptations demonstrate that inconsistency measures can be meaningfully extended beyond classical propositional logic when domain-specific semantics are carefully considered.

\textbf{Argumentation Frameworks:} Work on measuring inconsistency in argumentation frameworks has adapted traditional measures to handle attack relations between arguments rather than direct logical conflicts~\cite{Hunter2006}. This demonstrates that inconsistency measures can be extended to graph-based representations where conflicts arise from structural relationships.

\textbf{Non-Monotonic Logic:} \cite{Ulbricht2016} adapted inconsistency measures to Answer Set Programming, addressing the fundamental challenge that non-monotonic reasoning violates classical monotonicity postulates. Their work shows that adding information to a non-monotonic knowledge base can actually reduce inconsistency, requiring reformulation of rationality postulates for non-monotonic contexts.

\textbf{Temporal Logic:} \cite{Corea2022} developed inconsistency measures for temporal logic applied to declarative process specifications, where traditional measures fail to adequately capture temporal operators or distinguish conflicts affecting different numbers of future states. Their work shows how inconsistency measures can be adapted to handle temporal modalities beyond static propositional assertions.

These successful adaptations across argumentation, non-monotonic reasoning, and temporal logic establish that inconsistency measures can be meaningfully extended to diverse logical formalisms. Each adaptation required careful consideration of domain-specific conflict semantics, providing methodological precedent for adapting measures to classical planning.

\subsubsection{Computational Approaches and Non-Monotonic Considerations}

Planning applications require computational efficiency and compatibility with non-monotonic reasoning systems. Experimental comparisons demonstrate that \ac{ASP}-based algorithms often outperform SAT-based approaches for computing inconsistency measures~\cite{Kuhlmann2025}, making \ac{ASP} a promising foundation for planning applications.

The theoretical groundwork for \ac{ASP}-based planning was established by~\cite{Dimopoulos1997}, with modern tools like plasp 3~\cite{Dimopoulos2019} providing efficient translations from \ac{PDDL} to \ac{ASP}. Importantly, inconsistency measures for \ac{ASP} have been developed~\cite{Ulbricht2016}, with the crucial observation that the monotonicity postulate does not hold in non-monotonic logics.

This non-monotonic behavior is directly relevant for planning domains: adding information (such as additional actions or initial state facts) can actually reduce inconsistency by enabling new solution paths. This violates the classical monotonicity postulate and requires careful consideration when adapting inconsistency measures to planning contexts.

\subsection{Classical Planning}

Classical planning operates on factored representations, meaning the world state is decomposed into individual variables like ``robot at location A'' or ``door is open'' rather than treating each complete world situation as an indivisible atomic state~\cite{Russell2016}. A planning problem is formally defined as a 4-tuple $\langle P, O, I, G \rangle$ where $P$ is a finite set of atomic propositions, $O$ is a finite set of operators, $I \subseteq P$ is the initial state, and $G \subseteq P$ is the goal specification~\cite{Bylander1994}. Each state $s$ is a subset $s \subseteq P$ representing the propositions that are true in that state, following database semantics where the closed-world assumption treats unmentioned propositions as false and the unique names assumption ensures distinct objects are different~\cite{Russell2016}.

Actions are described by ground operators, which are fully instantiated actions with all variables replaced by specific objects, denoted as $o \in O$ and defined by the triple
\begin{equation*}
    \langle \text{precond}(o), \text{add}(o), \text{delete}(o) \rangle\,,
\end{equation*}
where $\text{precond}(o) \subseteq P$ are the preconditions, $\text{add}(o) \subseteq P$ are the add effects, and $\text{delete}(o) \subseteq P$ are the delete effects~\cite{Bylander1994}. An operator $o$ is applicable in state $s$ if and only if $\text{precond}(o) \subseteq s$. The result of executing action $o$ in state $s$ is defined as $\text{RESULT}(s,o) = (s \setminus \text{delete}(o)) \cup \text{add}(o)$~\cite{Russell2016}. A plan is a sequence of operators $\langle o_1, o_2, \ldots, o_k \rangle$ such that applying this sequence from the initial state $I$ results in a state $s$ where $G \subseteq s$~\cite{Bylander1994}.

A planning problem $\langle P, O, I, G \rangle$ is \emph{unsolvable} if there does not exist a sequence of operators $\pi = \langle o_1, o_2, \ldots, o_k \rangle$ such that each $o_i$ is applicable in the state resulting from applying $\langle o_1, \ldots, o_{i-1} \rangle$ to $I$, and the final state satisfies $G \subseteq s_k$. More precisely, defining the sequential application of operators as $s_0 = I$ and $s_i = \text{RESULT}(s_{i-1}, o_i)$ for $i = 1, \ldots, k$, a planning problem is unsolvable when no such sequence exists with $G \subseteq s_k$. Determining whether a plan exists is PSPACE-complete~\cite{Bylander1994}, which has motivated extensive research into efficient algorithms and search techniques.

The \ac{STRIPS} representation, originally developed by~\cite{Fikes1971}, is foundational for many modern planning systems, providing a simple yet powerful framework for modeling planning domains where actions have deterministic effects and the world is fully observable.

\ac{PDDL} serves as the standard language for describing planning problems and is used in international planning competitions~\cite{McDermott1998,Ghallab1998}. However, given the complexity of \ac{PDDL}'s various extensions and features, this thesis initially focuses on \ac{STRIPS}-based formulations due to their foundational simplicity and well-understood semantics, which allow for systematic adaptation of inconsistency measures without the complications of advanced \ac{PDDL} constructs.

Modern planning systems employ various search strategies to explore different ways of reaching goals, such as searching forward from the initial state, working backward from the goal, or examining the space of possible plans~\cite{Ghallab2004,Russell2016}. These strategies include techniques that build graphical representations of planning problems to identify conflicts and estimate solution difficulty~\cite{Blum1997}, and heuristic search methods~\cite{Bonet2001,Hoffmann2001} that use simplified versions of problems to guide their search toward solutions. These heuristic approaches have significantly improved planning performance for solvable instances, but when these systems encounter an unsolvable problem, they typically provide minimal diagnostic information beyond reporting failure~\cite{Eriksson2018}. This diagnostic gap represents a significant limitation for users attempting to understand the source of conflicts or determine appropriate modifications to achieve solvability.

\subsubsection{Taxonomy of Unsolvability Types}

Unsolvable planning problems can be classified by their structural cause. This taxonomy provides a framework for conflict analysis and guides measure adaptation to planning.

\textbf{Type 1: Goal Conflicts.} The goal specification $G$ contains propositions that cannot simultaneously be true in any valid state. The conflict exists in the goal specification itself, independent of the available actions or initial state. For instance, a goal requiring both $\{\text{light\_on}, \text{light\_off}\}$ when these represent mutually exclusive conditions cannot be satisfied by any state.

\textbf{Type 2: Precondition Unreachability.} Required preconditions for goal-achieving actions cannot be satisfied from the initial state. This category includes two subtypes: (2a) problems where no operator produces a required proposition, such as unlocking a door when no operator can produce the required key, and (2b) problems where preconditions form circular dependencies, such as requiring a security badge to be activated to enter a building, but needing to be inside the building to activate the badge.

\textbf{Type 3: Dead-End States.} The state space contains reachable states from which no sequence of operators can achieve the goal. These ``trap states'' represent irreversible transitions that prevent goal achievement. For example, if collecting an item from a room causes the door to lock permanently, and the goal requires items from multiple rooms, collecting the first item creates a dead-end state.

\textbf{Type 4: Resource Insufficiency.} Consumable resources are insufficient to satisfy goal requirements. This includes two subtypes: (4a) total resource deficit, where the cumulative consumption exceeds the available supply (such as requiring two fuel-consuming moves but having only one unit of fuel), and (4b) resource contention, where multiple goals compete for the same limited resource (such as two surfaces requiring painting but only one can of paint being available).

This taxonomy reveals that unsolvability stems from distinct structural properties: goal specification (Type 1), precondition dependencies (Type 2), state space topology (Type 3), and resource constraints (Type 4). Each type suggests different approaches to inconsistency quantification, such as counting conflicting goal pairs, analyzing precondition graphs for missing producers or cycles, measuring dead-end regions, or computing resource deficits.

To illustrate Type 1 (Goal Conflicts) with a formal specification, consider the planning problem $\langle P, O, I, G \rangle$ with $P = \{\text{light\_on}, \text{light\_off}\}$, $I = \{\text{light\_off}\}$, and $G = \{\text{light\_on}, \text{light\_off}\}$. The operator set $O = \{\text{turn\_on}, \text{turn\_off}\}$ contains
\begin{align*}
    \text{turn\_on}  & = \langle \{\text{light\_off}\}, \{\text{light\_on}\}, \{\text{light\_off}\} \rangle \\
    \text{turn\_off} & = \langle \{\text{light\_on}\}, \{\text{light\_off}\}, \{\text{light\_on}\} \rangle
\end{align*}
Since states are subsets of $P$ representing true propositions, and \textit{light\_on} and \textit{light\_off} represent mutually exclusive conditions, no state can contain both propositions simultaneously. Therefore, no sequence of operators can produce a state $s$ where $G \subseteq s$, making the problem unsolvable regardless of the available operators.

\subsection{Answer Set Programming for Planning}

\ac{ASP} provides a declarative approach to planning that has gained significant attention due to its expressiveness and computational efficiency for complex reasoning tasks~\cite{Dimopoulos1997}. In \ac{ASP}, planning problems are encoded as logic programs where answer sets correspond to valid plans, enabling the use of efficient SAT-based solvers to find solutions.

The modern tool plasp 3~\cite{Dimopoulos2019} provides efficient translations from \ac{PDDL} to \ac{ASP}, handling both sequential and parallel planning encodings. The system processes planning problems by first translating them into a standardized intermediate format and then generating \ac{ASP} rules that encode the planning semantics. Sequential encodings $S(i)$ search for plans of exactly $i$ steps, while parallel encodings like $S^\forall(i)$ and $S^\exists(i)$ allow concurrent action execution with appropriate serialization constraints.

For inconsistency measurement applications, \ac{ASP} offers several advantages over traditional planning approaches. First, the declarative nature of \ac{ASP} programs allows for systematic analysis of conflicts through unsatisfiable cores and related computational structures~\cite{Ulbricht2016}. Second, experimental evidence demonstrates that \ac{ASP}-based algorithms generally outperform SAT-based approaches for computing inconsistency measures, particularly for measures involving minimal inconsistent subsets~\cite{Kuhlmann2025}. Third, \ac{ASP} provides natural support for non-monotonic reasoning, which is essential for planning domains where the monotonicity postulate of classical inconsistency measures may not hold due to the frame problem and default assumptions about action effects~\cite{Ulbricht2016}.

\subsection{Challenges in Adapting Inconsistency Measures to Planning}

While inconsistency measures have been successfully adapted to temporal domains, planning presents unique challenges that distinguish it from static knowledge bases and temporal constraint problems. Unlike propositional knowledge bases where inconsistency arises from contradictory formulas, planning inconsistency emerges from the interaction between initial states, action sequences, and goal specifications within a dynamic system.

The fundamental challenge lies in the temporal and causal nature of planning. In classical inconsistency measurement, conflicts are typically identified among static formulas through direct logical contradiction. In planning, however, conflicts arise through complex interactions: action preconditions that cannot be satisfied given available resources, goal states that are mutually exclusive and cannot be achieved simultaneously, or action sequences that lead to dead-end states from which no goal can be reached. These conflicts manifest not as direct logical contradictions but as reachability problems in the state space defined by the planning domain.

Furthermore, planning domains involve the frame problem, where unmentioned propositions are assumed to persist unchanged unless explicitly affected by actions. This introduces default assumptions that complicate the direct application of traditional inconsistency measures, which assume explicit representation of all relevant information. The closed-world assumption in planning domains means that conflicts may arise from the absence of enabling conditions rather than the presence of contradictory assertions.

Adapting inconsistency measures to planning requires addressing how conflicts manifest in the context of action sequences, state transitions, and goal reachability. Unlike temporal constraint problems where constraints operate on continuous variables with explicit temporal intervals, planning involves discrete state transitions governed by preconditions and effects.

Of particular relevance to this thesis, paraconsistent approaches such as the contension measure face unique interpretation challenges in planning domains. In classical propositional logic, assigning the truth value $B$ (``both'') to a proposition represents a direct logical contradiction between formulas. However, in planning, states are defined as subsets of propositions where each proposition is either explicitly true (in the state) or false (not in the state) under the closed-world assumption. The notion of a proposition being ``both true and false'' does not have an immediate interpretation in this binary state representation. Potential adaptations might involve temporal indexing of propositions (considering $p$ at different time points), analyzing conflicts in the goal specification where mutually exclusive propositions are required, or examining propositions that play contradictory roles across different components of the planning problem (initial state, actions, goals). The development of such adaptations requires careful consideration of what constitutes a meaningful ``conflicting'' proposition in the context of planning problems.

\subsubsection{Limitations of Naive Propositional Encoding}

To investigate whether direct application of paraconsistent measures might work despite these conceptual challenges, an exploration was conducted encoding planning problems as propositional knowledge bases and computing the contension measure $I_c$. Five representative unsolvable planning problems spanning all four taxonomy types were systematically encoded: goal conflicts, unreachable preconditions, dead-end states, total resource deficit, and resource contention.

For each problem, a knowledge base $K$ was constructed by extracting goal propositions and relevant constraints from the planning structure. Different constraint patterns were used for each unsolvability type: goal conflicts as $\neg(p \land q)$, unreachable preconditions as $\text{goal} \to \text{precond}_1 \to \cdots \to \neg\text{precond}_n$, dead-end states as $\text{goal}_1 \to \text{effect}$ and $\text{goal}_2 \to \neg\text{effect}$, and resource constraints as $\text{goal} \to \text{resource\_amount}$ with $\neg\text{resource\_amount}$.

Computing $I_c$ for these encodings revealed a fundamental limitation: all five examples yielded $I_c = 1$, regardless of unsolvability type or structural complexity. This uniform value fails to distinguish between qualitatively different sources of unsolvability. Consider the light switch problem from the taxonomy example with knowledge base
\begin{equation*}
    K = \{\text{light\_on}, \text{light\_off}, \neg(\text{light\_on} \land \text{light\_off})\}\,.
\end{equation*}
The interpretation with $\upsilon(\text{light\_on}) = B$ and $\upsilon(\text{light\_off}) = T$ satisfies all three formulas. Since $B \land T = B$ and $\neg B = B$ in three-valued logic, the mutual exclusivity constraint is satisfied despite only one proposition receiving value $B$. This yields $I_c = 1$. The symmetric interpretation with values reversed also works, but $I_c$ reports only the minimum cardinality, losing information about conflict symmetry.

This negative result demonstrates that simple propositional encoding is insufficient for adapting paraconsistent inconsistency measures to planning. The interaction between three-valued logic semantics and constraint encodings collapses distinctions that the measure should ideally capture. This finding motivates the need for planning-specific adaptations that operate directly on planning structures rather than translating to propositional knowledge bases.
