\section{Implementation}
\label{sec:implementation}

This chapter presents the \ac{ASP}-based implementation of the adapted inconsistency measures from Section~\ref{sec:proposed_measures}. The system accepts \ac{PDDL} planning problems as input and, given a horizon parameter $h$, computes the full diagnostic profile
\begin{equation*}
    (\mathcal{I}^{\text{scope}}_{\text{UR}}(\Pi, h), \mathcal{I}^{\text{struct}}_{\text{UR}}(\Pi, h), \mathcal{I}^{\text{scope}}_{\text{MX}}(\Pi, h), \mathcal{I}^{\text{struct}}_{\text{MX}}(\Pi, h), \mathcal{I}^{\text{scope}}_{\text{GS}}(\Pi, h), \mathcal{I}^{\text{struct}}_{\text{GS}}(\Pi, h))
\end{equation*}
through a three-phase pipeline: \ac{PDDL} translation, \ac{ASP} state-space exploration via brave reasoning, and Python-based measure extraction. The horizon parameter $h$ directly corresponds to the formal parameter in the measure definitions (Section~\ref{sec:horizon_bounded}).

\subsection{System Architecture}
\label{sec:architecture}

The implementation follows a pipeline architecture with three distinct phases, each handled by a dedicated component.

\paragraph{Phase 1: PDDL Translation.}
The \ac{PDDL} domain and problem files are translated into \ac{ASP} facts using \textit{plasp}~3 by Dimopoulos et al.~\cite{Dimopoulos2019}, a standard tool from the Potassco \ac{ASP} ecosystem. The translation produces a lifted \ac{ASP} program where propositions are represented as boolean-valued variables and operators as parameterized actions. For benchmark domains that declare action costs using the \texttt{:functions} requirement, a preprocessing step strips cost-related elements, as these are irrelevant to inconsistency measure computation.

\paragraph{Phase 2: ASP Brave Reasoning.}
The translated \ac{ASP} program is loaded alongside a custom reachability encoding and solved using Clingo~5 by Gebser et al.~\cite{Gebser2012} with the \texttt{--enum-mode=brave} configuration. Under brave reasoning (Definition~\ref{defn:brave_cautious}), Clingo computes the union of atoms that are true in \emph{any} answer set, without enumerating individual answer sets. This mode is central to the implementation: different answer sets correspond to different operator execution traces, and the union across all answer sets captures the full scope of reachable states and achievable propositions. Dedicated witness atoms record whether goal pairs can coexist in a single state or be achieved in sequence; the absence of such a witness under brave reasoning constitutes proof that no execution trace satisfies the corresponding condition, which is the detection mechanism for mutex and sequencing conflicts.

\paragraph{Phase 3: Measure Extraction.}
A Python post-processing module collects the brave reasoning atoms and computes all six measure values through set operations. Unreachability is determined by comparing goal propositions against the set of truly reachable propositions. Mutex and sequencing conflicts are identified by the absence of the corresponding witness atoms: if brave reasoning, which considers all possible execution traces, never derives a witness, the conflict is guaranteed to hold. The details of the witness definitions and their correspondence to the formal measure definitions are given in Section~\ref{sec:witness_collection}.

The implementation is written in Python~3.12 with Clingo~5.6+ as the sole computational dependency. Docker containerization enables consistent execution regardless of the environment, and an automated test suite ensures correctness and facilitates regression testing during development. The codebase is structured to allow both library usage and batch processing of benchmark domains, with configurable timeouts to handle grounding complexity in larger problems. The complete source code, including all \ac{ASP} encodings, the Python measure extraction module, and the benchmark runner, is publicly available~\cite{Bunge2026}.

\subsection{PDDL Translation}
\label{sec:pddl_translation}

The input to the system is a pair of standard \ac{PDDL} files: a domain specification defining predicates and action schemas, and a problem specification defining objects, the initial state, and goal conditions.

\subsubsection{Initial Translator}

An initial implementation used the \texttt{pddl} Python library by Favorito~\cite{Favorito2021} to parse \ac{PDDL} files and ground action schemas through typed Cartesian product enumeration. For each action schema, all possible substitutions of typed objects were generated, and for each valid substitution, the grounded operator was emitted as \ac{ASP} facts. This approach was validated on all hand-crafted test scenarios from Section~\ref{sec:test_scenarios}, producing correct measure profiles for every case.

However, testing on the IPC~2016 Unsolvability Competition benchmarks by Muise~\cite{Muise2016} revealed two critical limitations. First, the \texttt{pddl} library does not support the \texttt{:constants} declaration used by several IPC domains, notably \textit{bag-gripper} and \textit{cave-diving}, causing translation failures. Second, the naive Cartesian product grounding strategy produces prohibitively large \ac{ASP} programs for domains that use predicates to constrain parameter interactions. For example, the \textit{pegsol} domain, where the \texttt{in-line} predicate constrains which positions can participate in a move, produced over 395{,}000 ground \ac{ASP} facts because the translator enumerated all position triples regardless of the \texttt{in-line} relationship.

\subsubsection{Adoption of plasp}
\label{sec:plasp_adoption}

To address these limitations, the implementation was restructured to use \textit{plasp}~3~\cite{Dimopoulos2019} for \ac{PDDL} translation. The \textit{plasp} tool, developed as part of the Potassco project, translates \ac{PDDL} specifications into a lifted \ac{ASP} representation where Clingo handles the grounding process. This offers two advantages over the initial approach: first, \textit{plasp} correctly handles \ac{PDDL} features including \texttt{:constants} and complex type hierarchies; second, Clingo's grounding algorithm exploits the structure of \ac{ASP} rules to avoid generating impossible parameter combinations.

The \textit{plasp} output format represents propositions as boolean-valued variables and operators as parameterized actions with compound terms. Table~\ref{tab:plasp_vocabulary} shows the correspondence between the internal representation used by the reachability encoding and the \textit{plasp} output format.

\begin{table}[ht]
    \centering
    \small
    \begin{tabular}{lll}
        \hline
        \textbf{Concept} & \textbf{plasp format}                              & \textbf{Internal format} \\
        \hline
        Initial state    & \texttt{initialState(V, value(V,true))}            & \texttt{init(P)}         \\
        Goal             & \texttt{goal(V, value(V,true))}                    & \texttt{goal(P)}         \\
        Precondition     & \texttt{precondition(A, V, value(V,true))}         & \texttt{precond(O, P)}   \\
        Add effect       & \texttt{postcondition(A,\,..., V, value(V,true))}  & \texttt{add(O, P)}       \\
        Delete effect    & \texttt{postcondition(A,\,..., V, value(V,false))} & \texttt{delete(O, P)}    \\
        \hline
    \end{tabular}
    \caption{Mapping between internal and \textit{plasp} predicates.}
    \label{tab:plasp_vocabulary}
\end{table}

A bridge encoding translates between these representations. Since \ac{ASP} distinguishes predicates by both name and arity, \textit{plasp}'s \texttt{goal/2} and the internal \texttt{goal/1} coexist without conflict.

\input{figures/lst_bridge}

The following example illustrates the complete translation for a small planning problem that exercises every bridge rule.

\begin{example}[Bridge Translation]
    \label{ex:bridge_translation}
    Consider a planning problem with two operators: \texttt{unlock} requires being at the door and the door being locked, then removes the lock; \texttt{enter} requires being at the door and the door \emph{not} being locked, then adds \texttt{in\_room} and removes \texttt{at\_door}. The \textit{plasp} output for this problem is:

    \input{figures/lst_plasp_example}

    Applying the bridge encoding from Listing~\ref{lst:bridge} yields the following internal representation:

    \input{figures/lst_bridge_result}

    Note how the \texttt{value(\ldots, false)} terms produce two semantically distinct internal predicates: \texttt{neg\_precond(enter, locked)} encodes that \texttt{enter} requires \texttt{locked} to be false before application, while \texttt{delete(unlock, locked)} encodes that \texttt{unlock} makes \texttt{locked} false as an effect. This distinction is central to the reachability encoding: negative preconditions restrict operator applicability (Section~\ref{sec:asp_encoding}), while delete effects drive state change and are the root cause of mutex and sequencing conflicts.
\end{example}

\subsubsection{PDDL Preprocessing}

Several IPC~2016 domains declare action costs through the \texttt{:functions} requirement, which \textit{plasp} does not support. Action costs affect only plan quality (the cost of a solution) and not plan existence (whether any solution exists); since the inconsistency measures defined in this thesis operate exclusively on reachability and achievability, removing cost annotations is semantically safe. A preprocessing step therefore removes cost-related elements: the \texttt{:action-costs} requirement, \texttt{(:functions ...)} declarations, \texttt{(increase ...)} and \texttt{(decrease ...)} effect terms, numeric function value assignments, and \texttt{(:metric ...)} specifications. After preprocessing, domains such as \textit{bag-transport} and \textit{over-nomystery} translate successfully.

Two IPC domains (\textit{bag-barman} and \textit{tetris}) use the \texttt{:equality} requirement with variable comparison expressions \texttt{(= ?x ?y)} in preconditions, which is not supported by \textit{plasp}. These domains remain incompatible and are excluded from the evaluation.

\subsection{ASP Encoding for Measure Computation}
\label{sec:asp_encoding}

The core of the implementation is a pair of \ac{ASP} encodings that derive proposition and operator sets from the translated facts and perform horizon-bounded state-space exploration with brave reasoning. Following standard practice in the \ac{ASP} literature~\cite{Lifschitz2019,Gebser2012}, the encodings are presented directly, as the declarative nature of \ac{ASP} rules makes them serve simultaneously as formal specification and executable program.

\subsubsection{Planning Representation}

The planning encoding derives the proposition and operator sets from the input facts:

\input{figures/lst_planning}

\subsubsection{State-Space Exploration}

The main computation performs horizon-bounded state-space exploration. The encoding models time-indexed state evolution where \texttt{holds(P, T)} indicates that proposition $P$ is true at time step $T$:

\input{figures/lst_exploration}

The choice rule on line~12 selects at most one applicable operator at each time step. Different selections produce different answer sets, each corresponding to a distinct execution trace through the state space. The frame axiom (line~17) ensures that propositions persist unless explicitly deleted.

\subsubsection{Brave Reasoning for Witness Collection}
\label{sec:witness_collection}

The key insight of the implementation is the use of Clingo's brave reasoning mode (\texttt{--enum-mode=brave}). Under this configuration, Clingo computes the union of atoms that are true in at least one answer set, without enumerating individual answer sets. This is essential for efficiency: the number of possible execution traces grows exponentially with the horizon, but the brave consequences can be computed without explicit enumeration.

The choice of brave over cautious reasoning is deliberate. Brave reasoning computes the existential union $\text{BRAVE}(\Pi)$ (Definition~\ref{defn:brave_cautious}), so an atom appears in the result if and only if at least one answer set contains it. For conflict detection, the converse is what matters. If a witness atom is \emph{absent} from the brave consequences, then no answer set contains it, which provides a universal proof that the corresponding condition is unsatisfiable across all execution traces. Cautious reasoning would yield atoms true in \emph{every} answer set, which is the wrong semantics for this purpose. A witness atom absent from the cautious consequences might still be derivable in some trace, meaning the corresponding conflict does not actually hold.

Three categories of witness atoms are collected:

\input{figures/lst_witnesses}

Under brave reasoning, \texttt{true\_reachable(P)} is derived if and only if proposition $P$ appears in at least one state of at least one execution trace. A proposition is truly unreachable only if no execution trace of any length up to the horizon can produce a state containing it.

The \texttt{coexist\_witness(G1, G2)} atom is derived if some execution trace reaches a state where both $G_1$ and $G_2$ hold simultaneously. The absence of this witness under brave reasoning means that no execution trace, across all possible operator selections, ever produces a state satisfying both goals. This corresponds exactly to the mutex relationship from Definition~\ref{defn:mutex_goal_pair}.

The \texttt{g2\_after\_g1\_witness(G1, G2)} atom is derived if some execution trace reaches $G_1$ at time $T_1$ and $G_2$ at some $T_2 \geq T_1$. Its absence means that achieving $G_1$ necessarily prevents subsequent achievement of $G_2$ in every possible continuation, corresponding to the sequencing conflict from Definition~\ref{defn:sequencing_conflict}.

\subsection{Measure Extraction}
\label{sec:measure_extraction}

After the brave reasoning pass completes, a Python module collects the shown atoms via a callback mechanism. The measure values are computed through set operations on the collected atoms.

\paragraph{Unreachability Profile.}
Let $G_{\text{reach}} = G \cap \{\, p \mid \texttt{true\_reachable}(p) \,\}$ denote the set of reachable goal propositions and $P_{\text{reach}} = \{\, p \mid \texttt{true\_reachable}(p) \,\}$ the set of all reachable propositions. The unreachability profile is:
\begin{equation*}
    \mathcal{I}^{\text{scope}}_{\text{UR}}(\Pi, h) = |G \setminus G_{\text{reach}}| \qquad \mathcal{I}^{\text{struct}}_{\text{UR}}(\Pi, h) = |P \setminus P_{\text{reach}}|
\end{equation*}

\paragraph{Goal Mutex Profile.}
For each unordered pair of achievable goals $(g_1, g_2) \in \binom{G_{\text{reach}}}{2}$, the pair is mutex if $\texttt{coexist\_witness}(g_1, g_2)$ was not derived under brave reasoning. The mutex profile counts the number of goals participating in at least one mutex pair and the total number of mutex pairs.

\paragraph{Goal Sequencing Profile.}
For each ordered pair of achievable goals $(g_1, g_2) \in G_{\text{reach}} \times G_{\text{reach}}$ with $g_1 \neq g_2$, the pair is a sequencing conflict if $\texttt{g2\_after\_g1\_witness}(g_1, g_2)$ was not derived. The sequencing profile counts the number of goals involved in at least one conflict and the total number of ordered conflict pairs.

The resulting six values are packaged into a diagnostic profile data structure that also provides automatic category classification according to the taxonomy from Section~\ref{sec:taxonomy}.

\subsection{Tooling and Reproducibility}
\label{sec:tooling}

The implementation provides both a library API and a command-line interface. The library exposes a single entry point, \texttt{compute\_measures}, that accepts either pre-translated \ac{ASP} files or \ac{PDDL} domain and problem files with a horizon parameter $h$. This function directly implements the parameterized measures $\mathcal{I}(\Pi, h)$ from Section~\ref{sec:proposed_measures}.

A batch processing module iterates over benchmark directories, automatically discovering domain and problem file pairs using IPC naming conventions (\texttt{domain.pddl} with \texttt{prob*.pddl}, or numbered \texttt{dom01.pddl} with \texttt{prob01.pddl}). Results are written to CSV format with problem size metadata (goal, proposition, and operator counts), per-problem timing, and error reporting.

The batch runner and command-line interface accept a configurable timeout parameter (in seconds). When the time limit is exceeded, the problem is recorded with \texttt{TIMEOUT} status rather than blocking indefinitely. This is necessary for benchmark domains where grounding complexity causes prohibitive solve times.

Implementing reliable timeouts for \ac{ASP} solving proved non-trivial. Clingo's \texttt{--time-limit} option is unavailable in the Python package distribution, and signal-based approaches (e.g., \texttt{SIGALRM}) are unreliable because Clingo's C extension can block signal delivery during grounding. The solution runs each computation in a separate child process using Python's \texttt{multiprocessing} module with the \texttt{spawn} start method, which creates a clean process that can be terminated with \texttt{SIGKILL} after the timeout, guaranteeing termination regardless of the solver's internal state.

For reproducibility, the implementation is containerized with Docker, bundling Python~3.12, Clingo~5.6, and \textit{plasp}~3 in a consistent environment. The Dockerfile and build instructions are included in the repository.

\subsection{Validation}
\label{sec:validation}

Correctness is verified through an automated test suite comprising 36 tests across four categories.

\paragraph{Scenario Tests.}
Eleven test scenarios are tested against their theoretically derived expected profiles. Table~\ref{tab:validation_results} shows the computed profiles alongside the expected values. All eleven scenarios produce exact matches, including the \textit{Negative Precondition} scenario that exercises the \texttt{neg\_precond} encoding path.

\begin{table}[ht]
    \centering
    \begin{tabular}{llcc}
        \hline
        \textbf{Category} & \textbf{Scenario}     & \textbf{Expected} & \textbf{Computed} \\
        \hline
        P1 Unreachability & Locked Door           & $(1,3,0,0,0,0)$   & $(1,3,0,0,0,0)$   \\
                          & Bank Vault            & $(1,7,0,0,0,0)$   & $(1,7,0,0,0,0)$   \\
        P2 Mutex          & Light Switch          & $(0,0,2,1,0,0)$   & $(0,0,2,1,0,0)$   \\
                          & Traffic Light         & $(0,0,3,3,0,0)$   & $(0,0,3,3,0,0)$   \\
        P3 Sequencing     & Trust and Travel      & $(0,0,2,1,2,1)$   & $(0,0,2,1,2,1)$   \\
                          & Rival Alliances       & $(0,0,2,1,2,2)$   & $(0,0,2,1,2,2)$   \\
        Edge Cases        & Coexisting Goals      & $(0,0,0,0,0,0)$   & $(0,0,0,0,0,0)$   \\
                          & Single Goal           & $(0,0,0,0,0,0)$   & $(0,0,0,0,0,0)$   \\
                          & Empty Goals           & $(0,0,0,0,0,0)$   & $(0,0,0,0,0,0)$   \\
                          & Delete Relaxation     & $(1,1,0,0,0,0)$   & $(1,1,0,0,0,0)$   \\
                          & Negative Precondition & $(1,1,0,0,0,0)$   & $(1,1,0,0,0,0)$   \\
        \hline
    \end{tabular}
    \caption{Computed profiles for the test scenarios.}
    \label{tab:validation_results}
\end{table}

\paragraph{Hierarchy Tests.}
Four tests verify the measure relationships established in Section~\ref{sec:measure_relationships}: unreachable goals do not participate in mutex or sequencing conflicts, sequencing conflicts imply mutex, mutex without sequencing indicates reversible conflicts, and delete effects correctly prevent false reachability claims.

\paragraph{API Contract Tests.}
Eight tests verify the public interface: five tests confirm that the \texttt{MeasureProfile} dataclass correctly classifies categories and formats output, and three tests verify that \texttt{compute\_measures} raises appropriate errors for missing files and unsatisfiable programs, and that insufficient horizon values correctly surface as false unreachability.

\paragraph{PDDL Pipeline Tests.}
Thirteen tests verify the end-to-end \ac{PDDL} pipeline: four tests confirm that the \textit{plasp} translation and bridge encoding produce correct measure profiles, three tests verify that the batch file discovery correctly identifies domain/problem pairs using IPC naming conventions, and six tests verify the cost-stripping preprocessor.

All 36 tests execute in under one second, enabling rapid regression testing during development.
