\section{Related Work}
\label{sec:related_work}

The connection between inconsistency and planning has been explored from several perspectives, with foundational work establishing important theoretical and diagnostic approaches for unsolvable planning problems. Thimm~\cite{Thimm2019} and Grant et al.~\cite{Grant2023} survey the broader field of inconsistency measurement in propositional logic, where various techniques quantify conflicts in logical knowledge bases. Their systematic adaptation to planning domains remains an open research direction. Recent work by Kuhlmann et al.~\cite{Kuhlmann2025a} applies inconsistency metrics to diagnose non-executable plans in robotics and cognitive systems, demonstrating the emerging interest in bridging these fields.

Important theoretical groundwork on proof systems for demonstrating unsolvability in planning has been established by Eriksson et al.~\cite{Eriksson2018}. This work introduces the concept of ``dead states,'' defined as states from which no goal state is reachable, and proposes a rule-based proof system for certifying unsolvability. This concept connects directly to the sequencing conflict measure proposed in this thesis: a sequencing conflict between goals $g_1$ and $g_2$ exists precisely when all reachable states satisfying $g_1$ are dead with respect to achieving $g_2$. While Eriksson et al.~\cite{Eriksson2018} provide binary certificates (solvable/unsolvable), the proposed measure quantifies the extent of such conflicts, revealing how many goal pairs exhibit sequencing conflicts and whether conflicts are unidirectional or symmetric.

From a diagnostic perspective, Göbelbecker et al.~\cite{Goebelbecker2010} introduced ``excuses'' for unsolvable planning tasks: minimal modifications required to restore solvability, such as adding initial facts, introducing operators, or weakening goals. An ``add producer'' excuse introduces an operator whose effects include a missing proposition, while an ``add reversal operator'' excuse introduces an operator that undoes an irreversible action. While excuses provide actionable repair suggestions, the measures proposed here complement this by providing quantitative structural information without prescribing specific repairs. The unreachability measure identifies which propositions cannot be achieved, suggesting where add-producer excuses are needed; the sequencing conflict measure identifies irreversible action sequences, suggesting where add-reversal excuses apply. Following the inconsistency measurement principle of ``measuring what \emph{is}, not what \emph{should be},'' the proposed measures report objective structural properties, leaving repair decisions to domain experts.

The planning literature has also extensively studied mutex (mutual exclusion) relationships, particularly in the context of Graphplan~\cite{Blum1997}. Classical mutex analysis identifies proposition pairs that cannot be simultaneously true at a given planning level, used heuristically during search. The mutex measure proposed in this thesis extends this concept to goal-level diagnostics, identifying goals that are each individually achievable but never co-occur in any reachable state, quantifying both scope (number of involved goals) and structure (number of mutex pairs).

Complementing these approaches, Sreedharan et al.~\cite{Sreedharan2019} developed methods for producing explanations that help users understand why planning tasks are unsolvable, focusing on the identification of unreachable subgoals. While valuable for understanding individual cases, these explanation methods do not provide the quantitative measures needed for systematic comparison across different unsolvable instances.

The adaptation of inconsistency measures to non-propositional domains provides important methodological precedents for this thesis. Ulbricht et al.~\cite{Ulbricht2016} adapted inconsistency measures to \ac{ASP}, demonstrating that non-monotonic reasoning requires reformulated rationality postulates. Their key observation, that adding information to a non-monotonic knowledge base can reduce inconsistency, directly parallels the planning context, where adding operators can restore solvability. This motivates the Operator Monotonicity postulate (Section~\ref{sec:postulates}), which reverses the direction of classical monotonicity.

Corea et al.~\cite{Corea2022} developed inconsistency measures for declarative process specifications expressed in temporal logic, addressing a domain with structural parallels to planning. Their work identifies limitations of direct propositional measure application when temporal operators and activity orderings are involved, similar to the finding in this thesis that naive propositional encoding of planning problems yields uninformative measure values (Section~\ref{sec:naive_encoding}). Both domains involve sequential constraint satisfaction over time, though planning operates on state transitions while process specifications operate on activity traces. Corea et al.'s approach of developing domain-specific semantics for conflict detection, rather than reducing to propositional logic, validates the strategy of defining planning-native measures that operate directly on goals, operators, and reachability.

More broadly, the inconsistency measurement literature has established that different domains require different notions of what constitutes a ``conflict.'' In propositional logic, conflicts are direct contradictions; in temporal logic, conflicts involve unsatisfiable constraint combinations over future states; in planning, conflicts manifest as unreachable propositions, mutex goal pairs, or irreversible sequencing. Each domain adaptation preserves the core methodology (formal definitions, rationality postulates, complexity analysis) while instantiating domain-appropriate semantics.

These research efforts establish important foundations for understanding unsolvable planning problems. However, they share a common limitation: existing approaches primarily focus on binary characterizations or prescriptive repairs rather than quantitative structural analysis. This thesis bridges the gap by adapting inconsistency measurement methodology to provide systematic quantitative diagnostics for planning problems.
