\subsubsection{Rationality Postulates}
\label{sec:postulates}

A core component of inconsistency measurement research is analyzing which rationality postulates measures satisfy~\cite{Hunter2008,Thimm2019}. As demonstrated by Ulbricht et al.~\cite{Ulbricht2016}, domain-specific adaptations may require modified postulates. Classical postulates cannot be applied directly to planning problems for two reasons: first, planning has fundamentally different structure than propositional knowledge bases, involving goals, operators, and states rather than formulas; second, planning exhibits inherently non-monotonic behavior, as adding operators can reduce inconsistency by enabling new solution paths.

\begin{defn}[Planning Consistency]
    An inconsistency measure $I$ for planning problems satisfies \emph{Planning Consistency} if:
    \begin{equation*}
        I(\langle P, O, I, G \rangle) = 0 \text{ iff } \langle P, O, I, G \rangle \text{ is solvable}
    \end{equation*}
\end{defn}

This adapts the classical Consistency postulate ($I(K) = 0$ iff $K$ is consistent) to the planning domain, where solvability replaces logical consistency.

\begin{defn}[Operator Monotonicity]
    An inconsistency measure $I$ satisfies \emph{Operator Monotonicity} if for all planning problems $\langle P, O, I, G \rangle$ and operators $o$:
    \begin{equation*}
        I(\langle P, O \cup \{o\}, I, G \rangle) \leq I(\langle P, O, I, G \rangle)
    \end{equation*}
\end{defn}

Note that Operator Monotonicity reverses classical monotonicity because adding operators can only enable new paths, potentially reducing inconsistency. This parallels the observation by Ulbricht et al.~\cite{Ulbricht2016} that non-monotonic logics require inverted monotonicity postulates.

\begin{defn}[Goal Monotonicity]
    An inconsistency measure $I$ satisfies \emph{Goal Monotonicity} if for all planning problems $\langle P, O, I, G \rangle$ and goals $g$:
    \begin{equation*}
        I(\langle P, O, I, G \cup \{g\} \rangle) \geq I(\langle P, O, I, G \rangle)
    \end{equation*}
\end{defn}

Goal Monotonicity preserves the direction of classical monotonicity: adding requirements (goals) can only increase or maintain inconsistency, never decrease it. The following propositions analyze postulate compliance for each measure.

\begin{prop}
    $I^{\text{scope}}_{\text{UR}}$ does not satisfy Planning Consistency.
\end{prop}
\begin{proof}
    Counter-example: The Light Switch problem has $I^{\text{scope}}_{\text{UR}} = 0$ (both goals \textit{on} and \textit{off} appear in reachable states) but is unsolvable (goals are mutex). Conversely, Planning Consistency requires $I = 0$ only for solvable problems.
\end{proof}

\begin{prop}
    $I^{\text{scope}}_{\text{UR}}$ satisfies Operator Monotonicity.
\end{prop}
\begin{proof}
    Adding an operator $o$ can only expand $S_{\text{reach}}$ or leave it unchanged. If goal $g$ was unreachable (not appearing in any state in $S_{\text{reach}}$), adding $o$ may make it reachable by enabling new state transitions. This can only reduce or maintain $I^{\text{scope}}_{\text{UR}}$, never increase it.
\end{proof}

\begin{prop}
    $I^{\text{scope}}_{\text{UR}}$ satisfies Goal Monotonicity.
\end{prop}
\begin{proof}
    Adding goal $g$ to $G$ results in exactly one of two cases: (a) $g$ appears in some state in $S_{\text{reach}}$, so $I^{\text{scope}}_{\text{UR}}$ is unchanged, or (b) $g$ does not appear in any state in $S_{\text{reach}}$, so $I^{\text{scope}}_{\text{UR}}$ increases by exactly 1.
\end{proof}

\begin{prop}
    $I^{\text{scope}}_{\text{MX}}$ does not satisfy Planning Consistency.
\end{prop}
\begin{proof}
    Counter-example: The Locked Door problem has $I^{\text{scope}}_{\text{MX}} = 0$ (no mutex relationships exist since the goal \textit{in\_room} is not even achievable) but is unsolvable.
\end{proof}

\begin{prop}
    $I^{\text{scope}}_{\text{MX}}$ does not satisfy Operator Monotonicity in general.
\end{prop}
\begin{proof}
    Adding an operator can introduce new reachable states, potentially creating new mutex relationships. Consider a problem where goal $g_2$ is initially unreachable (hence no mutex with $g_1$). Adding an operator that makes $g_2$ reachable but in states that never contain $g_1$ creates a new mutex pair, increasing $I^{\text{scope}}_{\text{MX}}$.
\end{proof}

\begin{prop}
    $I^{\text{scope}}_{\text{MX}}$ satisfies Goal Monotonicity.
\end{prop}
\begin{proof}
    Adding goal $g$ to $G$ can only increase the set of potential mutex pairs. If $g$ is mutex with existing goals, $I^{\text{scope}}_{\text{MX}}$ increases; otherwise it remains unchanged.
\end{proof}

\begin{prop}
    $I^{\text{scope}}_{\text{GS}}$ does not satisfy Planning Consistency.
\end{prop}
\begin{proof}
    Counter-example: The Light Switch problem has $I^{\text{scope}}_{\text{GS}} = 0$ (no sequencing conflicts since actions are reversible) but is unsolvable (goals are mutex).
\end{proof}

\begin{prop}
    $I^{\text{scope}}_{\text{GS}}$ does not satisfy Operator Monotonicity in general.
\end{prop}
\begin{proof}
    Counter-example: Consider $P = \{a, g_1, g_2\}$, $I = \{a\}$, $G = \{g_1, g_2\}$, and initial operator set $O = \{\text{reach\_g1}\}$ where $\text{reach\_g1}$ has $\text{precond} = \{a\}$, $\text{add} = \{g_1\}$, $\text{delete} = \{a\}$. Goal $g_1$ is achievable but $g_2$ is not, so no sequencing conflict exists. Adding operator $\text{reach\_g2}$ with $\text{precond} = \{a\}$, $\text{add} = \{g_2\}$, $\text{delete} = \{a\}$ makes $g_2$ achievable. However, from state $\{g_1\}$, neither operator is applicable since both require $\{a\}$. Thus $(g_1, g_2)$ becomes a sequencing conflict, increasing $I^{\text{scope}}_{\text{GS}}$ from 0 to 2.
\end{proof}

\begin{prop}
    $I^{\text{scope}}_{\text{GS}}$ satisfies Goal Monotonicity.
\end{prop}
\begin{proof}
    Adding goal $g$ to $G$ can only increase the set of potential sequencing conflicts. New conflicts of the form $(g, g')$ or $(g', g)$ may emerge; existing conflicts are unaffected.
\end{proof}

The structure variants $I^{\text{struct}}_{\text{UR}}$, $I^{\text{struct}}_{\text{MX}}$, and $I^{\text{struct}}_{\text{GS}}$ satisfy the same postulates as their scope counterparts. For Operator Monotonicity, only $I^{\text{struct}}_{\text{UR}}$ satisfies the postulate: adding operators expands reachability, reducing unreachable proposition counts. Both Mutex and Sequencing structure variants inherit their scope variants' violations: an operator making a previously unreachable goal reachable can introduce new mutex pairs or sequencing conflicts if the newly reachable goal states are incompatible with existing goal states. For Goal Monotonicity: adding goals increases the domain over which unreachable propositions, mutex pairs, and sequencing conflicts are counted.

Beyond postulate compliance, the following proposition establishes how the three measures relate to each other, forming a diagnostic hierarchy rather than independent failure detectors.

\begin{prop}[Measure Hierarchy]
    \label{prop:measure_hierarchy}
    The proposed measures form a diagnostic hierarchy:
    \begin{enumerate}
        \item If goal $g$ is unreachable, then $g$ does not participate in any
              mutex relationship or sequencing conflict
        \item If $(g_1, g_2)$ is a sequencing conflict (with $g_2$ achievable),
              then $g_1$ and $g_2$ are mutex
        \item Goals can be mutex without having any sequencing conflict
    \end{enumerate}
\end{prop}
\begin{proof}
    \begin{enumerate}
        \item Both mutex (Definition~\ref{defn:mutex_goal_pair}) and sequencing conflict
              (Definition~\ref{defn:sequencing_conflict}) require participating goals to be achievable.
              If $g$ is unreachable, it fails this condition.

        \item Suppose $(g_1, g_2)$ is a sequencing conflict with $g_2$ achievable.
              By Definition~\ref{defn:sequencing_conflict}, $g_1$ is achievable and for all $s \in S_{\text{reach}}$
              with $g_1 \in s$: $\nexists s' \in \text{REACH}(s) : g_2 \in s'$.
              Since $s \in \text{REACH}(s)$ (reflexivity), this implies $g_2 \notin s$ for all
              $g_1$-states. Thus no reachable state contains both $g_1$ and $g_2$. Combined with
              both goals being achievable, this satisfies Definition~\ref{defn:mutex_goal_pair}.

        \item Light Switch: Goals $\{\text{on}, \text{off}\}$ are mutex
              (both achievable, never co-occur) but have no sequencing conflict (from any
              \textit{on}-state, \textit{off} is reachable via \textit{turn\_off}, and vice versa).
    \end{enumerate}
\end{proof}

This hierarchy reflects increasing diagnostic specificity rather than
independent failure modes:
\begin{itemize}
    \item \textbf{Mutex without sequencing} (e.g., Light Switch): Goals exclude
          each other but actions are reversible. The conflict is symmetric;
          neither goal ``causes'' the problem more than the other.
    \item \textbf{Mutex with sequencing} (e.g., Trust and Travel): Goals exclude
          each other \emph{and} achieving specific goals creates permanent dead-ends.
          The sequencing measure identifies which goal achievement causes the
          blockage, providing actionable diagnostic information for adding
          reversal operators.
\end{itemize}

The analysis distinction remains meaningful: state space analysis (P1, P2)
determines what states exist, while trajectory analysis (P3)
determines what paths exist between states. For irreversible problems,
all three measures report non-zero values, with P3 providing directional
information about which goal achievements cause the blockage.

\begin{prop}[Scope-Structure Bounds]
    \label{prop:scope_structure_bounds}
    For any planning problem $\langle P, O, I, G \rangle$:
    \begin{enumerate}
        \item $I^{\text{scope}}_{\text{UR}} \leq I^{\text{struct}}_{\text{UR}} \leq |P|$
        \item $I^{\text{struct}}_{\text{MX}} \leq \binom{I^{\text{scope}}_{\text{MX}}}{2}$
        \item $I^{\text{struct}}_{\text{GS}} \leq I^{\text{scope}}_{\text{GS}} \times (I^{\text{scope}}_{\text{GS}} - 1)$
    \end{enumerate}
\end{prop}
\begin{proof}
    \begin{enumerate}
        \item Unreachable goals are a subset of all unreachable propositions,
              which are a subset of $P$.

        \item The maximum number of unordered pairs among
              $I^{\text{scope}}_{\text{MX}}$ goals is $\binom{I^{\text{scope}}_{\text{MX}}}{2}$.

        \item The maximum number of ordered pairs among
              $I^{\text{scope}}_{\text{GS}}$ goals is $n(n-1)$ where
              $n = I^{\text{scope}}_{\text{GS}}$.
    \end{enumerate}
\end{proof}

These bounds enable diagnostic interpretation through ratio analysis:

\begin{itemize}
    \item \textbf{Unreachability depth}: The ratio $(I^{\text{struct}}_{\text{UR}} - I^{\text{scope}}_{\text{UR}}) / I^{\text{scope}}_{\text{UR}}$ indicates cascading dependency depth. A ratio of 0 means goals are directly unreachable (no intermediate propositions blocked); larger ratios indicate deeper precondition chains.

    \item \textbf{Mutex density}: If $I^{\text{struct}}_{\text{MX}} = \binom{I^{\text{scope}}_{\text{MX}}}{2}$, the mutex goals form a complete clique where every pair is mutually exclusive, as in Traffic Light. Ratios below 1 indicate sparser conflict structures with isolated mutex pairs.

    \item \textbf{Sequencing symmetry}: If $I^{\text{struct}}_{\text{GS}} = I^{\text{scope}}_{\text{GS}} \times (I^{\text{scope}}_{\text{GS}} - 1)$, all sequencing conflicts are bidirectional: achieving any involved goal blocks all others, and vice versa. Ratios near 0.5 suggest predominantly unidirectional conflicts; lower ratios indicate star-shaped conflict patterns where one goal blocks many others but not conversely.
\end{itemize}

Table~\ref{tab:postulates} summarizes postulate compliance for the proposed measures.

\begin{table}[ht]
    \centering
    \caption{Postulate compliance of proposed planning inconsistency measures}
    \label{tab:postulates}
    \begin{tabular}{l|cc|cc|cc}
        \hline
                              & \multicolumn{2}{c|}{$I_{\text{UR}}$} & \multicolumn{2}{c|}{$I_{\text{MX}}$} & \multicolumn{2}{c}{$I_{\text{GS}}$}                                        \\
        Postulate             & scope                                & struct                               & scope                               & struct     & scope      & struct     \\
        \hline
        Planning Consistency  & \texttimes                           & \texttimes                           & \texttimes                          & \texttimes & \texttimes & \texttimes \\
        Operator Monotonicity & \checkmark                           & \checkmark                           & \texttimes                          & \texttimes & \texttimes & \texttimes \\
        Goal Monotonicity     & \checkmark                           & \checkmark                           & \checkmark                          & \checkmark & \checkmark & \checkmark \\
        \hline
    \end{tabular}
\end{table}

Table~\ref{tab:measure_values} demonstrates the diagnostic hierarchy across the test scenarios: unreachability problems show only $I_{\text{UR}}$, reversible mutex problems show only $I_{\text{MX}}$, while irreversible problems show both $I_{\text{MX}}$ and $I_{\text{GS}}$, consistent with Proposition~\ref{prop:measure_hierarchy}.

\begin{table}[ht]
    \centering
    \caption{Measure values across test scenarios (scope, structure)}
    \label{tab:measure_values}
    \begin{tabular}{l|cc|cc|cc}
        \hline
        Scenario         & \multicolumn{2}{c|}{$I_{\text{UR}}$} & \multicolumn{2}{c|}{$I_{\text{MX}}$} & \multicolumn{2}{c}{$I_{\text{GS}}$}                           \\
                         & scope                                & struct                               & scope                               & struct & scope & struct \\
        \hline
        Locked Door      & 1                                    & 3                                    & 0                                   & 0      & 0     & 0      \\
        Bank Vault       & 1                                    & 7                                    & 0                                   & 0      & 0     & 0      \\
        Light Switch     & 0                                    & 0                                    & 2                                   & 1      & 0     & 0      \\
        Traffic Light    & 0                                    & 0                                    & 3                                   & 3      & 0     & 0      \\
        Trust and Travel & 0                                    & 0                                    & 2                                   & 1      & 2     & 1      \\
        Rival Alliances  & 0                                    & 0                                    & 2                                   & 1      & 2     & 2      \\
        \hline
    \end{tabular}
\end{table}

The complete diagnostic profile for a planning problem is the 6-tuple:
\begin{equation*}
    \text{Profile} = (I^{\text{scope}}_{\text{UR}}, I^{\text{struct}}_{\text{UR}}, I^{\text{scope}}_{\text{MX}}, I^{\text{struct}}_{\text{MX}}, I^{\text{scope}}_{\text{GS}}, I^{\text{struct}}_{\text{GS}})
\end{equation*}

Each scenario in Table~\ref{tab:measure_values} exhibits a characteristic profile signature:
\begin{itemize}
    \item \textbf{Unreachability}: $(1, 3, 0, 0, 0, 0)$ for Locked Door indicates dependency chain failure with depth 2 (three unreachable propositions blocking one goal)
    \item \textbf{Reversible mutex}: $(0, 0, 2, 1, 0, 0)$ for Light Switch indicates a single mutex pair involving both goals; $(0, 0, 3, 3, 0, 0)$ for Traffic Light indicates a complete mutex clique (all $\binom{3}{2} = 3$ pairs are mutex)
    \item \textbf{Irreversible mutex}: $(0, 0, 2, 1, 2, 1)$ for Trust and Travel shows both mutex and unidirectional sequencing; $(0, 0, 2, 1, 2, 2)$ for Rival Alliances shows mutex with bidirectional sequencing
\end{itemize}

The sequencing structure value distinguishes unidirectional conflicts ($I^{\text{struct}}_{\text{GS}} = 1$ for Trust and Travel) from bidirectional conflicts ($I^{\text{struct}}_{\text{GS}} = 2$ for Rival Alliances), while the mutex structure value is identical ($I^{\text{struct}}_{\text{MX}} = 1$) for both. This illustrates how trajectory analysis (P3) provides finer-grained diagnostic information beyond state space analysis (P1, P2).

The observation that no single measure satisfies Planning Consistency motivates the multi-faceted diagnostic approach adopted in this work. As established in Proposition~\ref{prop:measure_hierarchy}, the measures form a diagnostic hierarchy with increasing specificity: unreachability precludes other conflict types, sequencing implies mutex, and mutex without sequencing indicates reversible conflicts. Scope-structure bounds (Proposition~\ref{prop:scope_structure_bounds}) enable diagnostic interpretation of conflict density and symmetry. This design parallels findings in~\cite{Ulbricht2016}, where multiple measures with different postulate profiles were necessary to adequately characterize inconsistency in answer set programs.
