\subsection{Structural Taxonomy for Unsolvable Planning Problems}
\label{sec:taxonomy}

Before presenting the proposed measures, this section establishes a structural taxonomy (Figure~\ref{fig:taxonomy}) that organizes unsolvable planning problems by features relevant to inconsistency measurement. This taxonomy identifies distinct measurement targets and motivates the design choices in subsequent sections.

\input{figures/fig_taxonomy}

\subsubsection{Structural Categories}

Unsolvable planning problems can be organized by structural features that suggest different approaches for inconsistency measurement. The taxonomy is organized by the analytical approach needed to detect and quantify each kind of conflict. The categories overlap rather than partition, and a single planning problem may exhibit features from more than one. The three structural categories are:

\textbf{Category 1: Operator-Induced Goal Conflicts.} The goal specification $G$ contains propositions that cannot all be true in any valid state. In classical propositional logic, this would include explicit contradictions such as $\{p, \neg p\}$. However, in propositional \ac{PDDL} with \texttt{:strips} requirements, goal specifications consist exclusively of positive literals. Negative goal literals require \ac{PDDL} 3.0 extensions~\cite{Gerevini2005}. Consequently, specification-level contradictions of the form $\{p, \neg p\}$ cannot occur in the planning formalism addressed by this thesis.

Nevertheless, goal propositions may be \emph{operator-induced mutex} (mutual exclusion). Each proposition is individually achievable, but action effects structurally prevent them from holding in the same reachable state. For instance, a goal requiring both $\{\textit{light\_on}, \textit{light\_off}\}$ when operators enforce mutual exclusion, where the turn-on operator deletes \textit{light\_off} and turn-off deletes \textit{light\_on}, cannot be satisfied by any reachable state. This operator-induced mutex represents the planning analog of specification-level conflicts. This category is detectable through reachability analysis of $G$ against the state space induced by $O$, and suggests measures that count goal propositions involved in mutex relationships and quantify the density of those relationships across the goal set.

\textbf{Category 2: Dependency and Effect Structure.} Problems arise in the precondition, add-effect, and delete-effect relationships between propositions and operators. The \emph{precondition-effect dependency graph} captures preconditions and add effects. Nodes represent propositions, and directed edges connect preconditions to effects through operators, indicating that achieving a proposition $p$ requires first satisfying proposition $q$. For example, if operator $o$ has $\text{precond}(o) = \{q\}$ and $\text{add}(o) = \{p\}$, the graph contains an edge from $q$ to $p$ labeled with $o$. Figure~\ref{fig:dependency_graph} illustrates this for the Locked Door scenario (Scenario~1 in Section~\ref{sec:test_scenarios}), where the proposition \textit{has\_key} has no producer, making it the root cause of unreachability. Delete effects are not captured by this graph and are addressed separately in subtype 2c.

\input{figures/fig_dependency_graph}

This category encompasses three structural features. The first two are analyzable via graph algorithms on the precondition-effect dependency graph, while the third extends to delete effects and requires trajectory analysis of how operator applications irreversibly alter reachable states:
\begin{itemize}
    \item \textbf{(2a) Missing producers:} No operator produces a required proposition. In the dependency graph, a goal proposition has no incoming edges. For example, unlocking a door when no operator can produce the required key.
    \item \textbf{(2b) Circular dependencies:} Preconditions form cycles without initialization. In the dependency graph, a cycle exists where no node in the cycle is initially true. For example, requiring a security badge to be activated to enter a building while needing to be inside the building to activate the badge.
    \item \textbf{(2c) Action-based exclusions:} Achieving certain goal propositions via irreversible actions creates dead-end states preventing other goal propositions from being reached. For example, entering a one-way door to retrieve an item while becoming trapped and unable to complete other objectives.
\end{itemize}
This category suggests reachability-based measures that count blocked goal propositions and trace their cascading dependency failures. Cycle detection and producer counting are natural extensions left to future work.

\textbf{Category 3: Resource Constraints.} Consumable resources such as time, fuel, budget, or battery are insufficient to satisfy goal requirements. This includes both total resource deficit, where the cumulative consumption exceeds the available supply, as when two fuel-consuming moves are required but only one unit of fuel is available, and resource contention, where multiple goal propositions compete for the same limited resource, as when two surfaces require painting but only one can of paint is available. This category requires numeric tracking and accounting measures, such as computing absolute or relative resource deficits, counting over-subscribed resources, or tracking path-dependent consumption patterns.

As an example of category overlap, resource contention problems in Category~3 often exhibit operator-induced goal conflicts from Category~1 caused by resource consumption. The taxonomy is organized by what measurement approaches are applicable: logical analysis for operator-induced goal conflicts, graph and trajectory analysis for dependency and effect structure, and numeric accounting for resources.

\subsubsection{Scope of This Thesis}

This work develops inconsistency measures for operator-induced goal conflicts in Category~1 and selected aspects of Category~2, focusing on conflicts detectable \linebreak through reachability-based analysis of propositional \ac{PDDL} planning problems. Spe\-cifically, three measures are proposed: the Goal Mutex Profile for operator-induced mutex relationships where goals are individually achievable but never satisfiable in a common reachable state, the Goal Sequencing Conflict Profile for irreversible action-based exclusions where achieving certain goals permanently blocks others, and the Unreachability Profile for generic unreachability diagnosis applicable across Category~2 subtypes.

Category~2b, circular dependencies, is recognized as an important source of unsolvability but is excluded from formal measurement in this thesis. Circular dependencies cause unreachability symptoms that are captured by the generic unreachability measure. Providing specific diagnostics for cycles, however, would require graph-theoretic analysis beyond reachability computation. While such analysis could yield more targeted repair suggestions, such as identifying which dependency edge to break versus which producer to add, this thesis prioritizes breadth of coverage over depth of cause-specific diagnosis. For instance, if no operator produces proposition \textit{has\_key}, a missing-producer case, the targeted repair is to add such an operator. By contrast, if the operator \textit{activate\_badge} has precondition \textit{inside} and the operator \textit{enter\_building} has precondition \textit{active\_badge}, forming a circular dependency, the targeted repair is either to break one dependency edge by modifying an operator's preconditions, or to initialize one proposition in the cycle as initially true. The generic unreachability measure identifies which propositions are blocked. Distinguishing why they are blocked, whether missing producer, cycle, or action-based exclusion, is recognized as valuable future work but is orthogonal to the current goal-level and dependency-level framework.

Category~3, resource constraints, is also excluded from formal measurement, for two reasons. First, from a knowledge representation perspective, numeric resource insufficiency often reduces to missing-producer problems in Category~2a. Second, numeric fluents introduce significant complexity through \ac{PDDL} 2.1 extensions~\cite{Fox2003}, requiring fundamentally different measurement approaches. This thesis focuses on propositional \ac{PDDL} planning with the \texttt{:strips} and \texttt{:typing} requirements. Numeric extensions warrant separate treatment. Future work may develop specialized measures for cycles and numeric resources, building on the foundational framework established here.

An important practical distinction emerges between static and dynamic analysis approaches. Static analyses operate directly on the problem specification $\langle P, O, I, G \rangle$ without state space exploration. Missing producers in Category~2a are detected by verifying whether $\exists o \in O$ with $p \in \text{add}(o)$. Operator-induced mutex in Category~1 admits a static over-approximation from delete-effect interactions, though exact detection requires reasoning about reachable states. Dynamic analyses, conversely, reason about reachable states or execution paths. Circular dependencies in Category~2b require graph analysis of precondition chains, exact unreachability across Category~2 requires forward or backward reachability computation, and irreversible action-based exclusions in Category~2c require trajectory analysis. This distinction has implications for computational complexity. Measures grounded in static over-approximation are typically more efficiently computable than those requiring exact dynamic state-space reasoning, motivating the bounded reachability approach adopted by the proposed measures.

\begin{example}[Multiple Structural Categories]
    To illustrate how multiple structural categories can manifest in a single planning problem, consider a security facility scenario:
    \begin{align*}
        P & = \{\text{outside}, \text{inside}, \text{has\_badge}, \text{vault\_accessed}, \text{lights\_on}, \text{lights\_off}\}       \\
        O & = \{ \text{enter\_facility} = \langle \{\text{outside}, \text{has\_badge}\}, \{\text{inside}\}, \{\text{outside}\} \rangle, \\
          & \quad \text{access\_vault} = \langle \{\text{inside}\}, \{\text{vault\_accessed}\}, \{\} \rangle,                           \\
          & \quad \text{turn\_on} = \langle \{\text{lights\_off}\}, \{\text{lights\_on}\}, \{\text{lights\_off}\} \rangle,              \\
          & \quad \text{turn\_off} = \langle \{\text{lights\_on}\}, \{\text{lights\_off}\}, \{\text{lights\_on}\} \rangle \}            \\
        I & = \{\text{outside}, \text{lights\_off}\}                                                                                    \\
        G & = \{\text{vault\_accessed}, \text{outside}, \text{lights\_on}, \text{lights\_off}\}
    \end{align*}

    This problem exhibits three distinct structural sources of unsolvability. First is Category~1, operator-induced mutex. The goal requires both \textit{lights\_on} and \textit{lights\_off}, but these propositions are mutually exclusive through operator effects, as achieving one necessarily deletes the other. Second is Category~2a, missing producer. The proposition \textit{has\_badge} appears in the precondition of \textit{enter\_facility} but no operator in $O$ has \textit{has\_badge} in its add effects, making facility entry impossible regardless of other constraints. Third is Category~2c, action-based exclusion. Even if a badge-producing operator existed, \textit{enter\_facility} deletes \textit{outside} without any operator to restore it, while the goal requires both \textit{vault\_accessed}, achievable only from inside, and \textit{outside}, creating an irreversible conflict. The three categories illustrate structurally distinct measurement targets within a single problem. Category~1 and Category~2a each independently render the problem unsolvable, while Category~2c is present as a structural pattern that would persist even if a producer for \textit{has\_badge} were added. This demonstrates that the taxonomy identifies measurement targets rather than only top-level causes of unsolvability.
\end{example}
