\subsubsection{Computational Complexity}
\label{sec:complexity}

This section analyzes the computational complexity of the proposed measures in terms of standard complexity classes. Determining plan existence, \textsc{PlanSat}, for propositional \ac{STRIPS} is PSPACE-complete~\cite{Bylander1994}. Standard complexity class definitions and the results cited below can be found in Papadimitriou~\cite{Papadimitriou1994}. All three measures depend on reachability analysis over the planning state space, and this structural connection to \textsc{PlanSat} determines their decision-problem complexity.

Following Thimm and Wallner~\cite{Thimm2016a} and Corea et al.~\cite{Corea2022}, four computational problems are defined for each measure $\mathcal{I} \in \{\mathcal{I}_{\text{UR}}, \mathcal{I}_{\text{MX}}, \mathcal{I}_{\text{GS}}\}$.

\begin{defn}[Decision and Function Problems]
    \label{defn:decision_problems}
    Let $\Pi = \langle P, O, I, G \rangle$ be a planning problem and let $k \in \mathbb{N}_0$. The following problems are defined for each measure $\mathcal{I}$:
    \begin{enumerate}
        \item $\text{UPPER}_{\mathcal{I}}$: is $\mathcal{I}^{\text{scope}}(\Pi) \leq k$?
        \item $\text{LOWER}_{\mathcal{I}}$: is $\mathcal{I}^{\text{scope}}(\Pi) \geq k$?
        \item $\text{EXACT}_{\mathcal{I}}$: is $\mathcal{I}^{\text{scope}}(\Pi) = k$?
        \item $\text{VALUE}_{\mathcal{I}}$: compute $\mathcal{I}^{\text{scope}}(\Pi)$.
    \end{enumerate}
\end{defn}

Analogous problems are defined for $\mathcal{I}^{\text{struct}}$. The scope variants iterate over $G$ or $G \times G$, while the structure variants iterate over $P$, unordered pairs from $G$, or ordered pairs from $G$, respectively. Since $|G| \leq |P|$ and all iteration domains are polynomial in the input size, scope and structure variants have identical decision-problem complexity. The following proofs are stated for the scope variants. The structure variants follow by the same membership arguments with the respective iteration domain, and hardness transfers since the scope problem reduces to any variant by the same polynomial composition.

Each completeness result below is established in two parts: \emph{membership} shows that the problem can be solved within the resource bounds of the complexity class, here polynomial space, while \emph{hardness} shows that the problem is at least as difficult as every other problem in that class, via reduction from a known complete problem. Together, these establish that the problem is exactly as hard as the hardest problems in the class.

\begin{thm}
    \label{thm:ur_complexity}
    $\text{UPPER}_{\mathcal{I}_{\text{UR}}}$ is PSPACE-complete.
\end{thm}
\begin{proof}
    \emph{Membership.} Checking whether $\mathcal{I}^{\text{scope}}_{\text{UR}} \leq k$ is equivalent to verifying that at least $|G| - k$ goal propositions are reachable. For each $g \in G$, the question ``does there exist a reachable state $s$ with $g \in s$?'' can be decided in NPSPACE: nondeterministically guess a sequence of operator applications starting from $I$, maintaining only the current state in polynomial space memory, and accept if a state containing $g$ is reached within $2^{|P|}$ steps. By Savitch's theorem~\cite{Papadimitriou1994}, NPSPACE $=$ PSPACE. Intuitively, space is reusable: nondeterministic branches can be explored sequentially in the same memory, incurring only a quadratic blowup that remains polynomial. Since $|G| \leq |P|$, iterating this check over all goals adds only polynomial overhead, and PSPACE is closed under polynomial time composition. Therefore $\text{UPPER}_{\mathcal{I}_{\text{UR}}} \in \text{PSPACE}$.

    \emph{Hardness.} We reduce from \textsc{PlanSat}. Given a \textsc{PlanSat} instance $\Pi = \langle P, O, I, G \rangle$, construct $\Pi' = \langle P', O', I, G' \rangle$ where:
    \begin{itemize}
        \item $P' = P \cup \{g^*\}$ for a fresh proposition $g^*$,
        \item $G' = \{g^*\}$,
        \item $O' = O \cup \{o^*\}$ where $o^* = \langle G, \{g^*\}, \emptyset \rangle$.
    \end{itemize}
    The operator $o^*$ is applicable exactly when all original goals are satisfied, producing $g^*$. Thus $g^*$ is reachable in $\Pi'$ if and only if $\Pi$ has a plan. Therefore $\mathcal{I}^{\text{scope}}_{\text{UR}}(\Pi') = 0$ if and only if $\Pi$ is solvable. Deciding $\text{UPPER}_{\mathcal{I}_{\text{UR}}}$ with $k = 0$ solves \textsc{PlanSat}, establishing PSPACE-hardness.
\end{proof}

\begin{cor}
    \label{cor:ur_all}
    $\text{LOWER}_{\mathcal{I}_{\text{UR}}}$ and $\text{EXACT}_{\mathcal{I}_{\text{UR}}}$ are PSPACE-complete. $\text{VALUE}_{\mathcal{I}_{\text{UR}}}$ is in FPSPACE.
\end{cor}
\begin{proof}
    $\text{LOWER}_{\mathcal{I}_{\text{UR}}}$ is the complement of $\text{UPPER}_{\mathcal{I}_{\text{UR}}}$ with threshold $k - 1$. Since PSPACE $=$ co-PSPACE~\cite{Papadimitriou1994}, it is also PSPACE-complete. This equality holds because PSPACE is a deterministic class: a deterministic polynomial space machine that decides membership in a language also decides its complement by simply negating the output. In the propositional setting of Thimm and Wallner~\cite{Thimm2016a}, the analogous relationship yields coNP-completeness from NP-completeness, a meaningful distinction. At the PSPACE level this distinction collapses.

    $\text{EXACT}_{\mathcal{I}_{\text{UR}}}$ can be decided by checking both $\mathcal{I}^{\text{scope}}_{\text{UR}} \leq k$ and $\mathcal{I}^{\text{scope}}_{\text{UR}} \geq k$. In the propositional setting, this conjunction of an NP problem and a coNP problem yields the class $D^p$, which contains but is believed to be strictly larger than $\text{NP} \cap \text{coNP}$. Here, the conjunction of two PSPACE problems remains in PSPACE. Hardness follows from the same reduction as Theorem~\ref{thm:ur_complexity}: since $\mathcal{I}^{\text{scope}}_{\text{UR}}(\Pi') \in \{0, 1\}$, deciding $\text{EXACT}_{\mathcal{I}_{\text{UR}}}$ with $k = 0$ solves \textsc{PlanSat}. Therefore PSPACE-complete.

    $\text{VALUE}_{\mathcal{I}_{\text{UR}}}$ can be computed via binary search with $O(\log |G|)$ calls to $\text{UPPER}_{\mathcal{I}_{\text{UR}}}$. Polynomially many PSPACE oracle calls remain in FPSPACE.
\end{proof}

\begin{thm}
    \label{thm:mx_complexity}
    $\text{UPPER}_{\mathcal{I}_{\text{MX}}}$ is PSPACE-complete.
\end{thm}
\begin{proof}
    \emph{Membership.} Checking whether $\mathcal{I}^{\text{scope}}_{\text{MX}} \leq k$ requires verifying that at most $k$ goals participate in mutex relationships. For each pair $(g_1, g_2) \in G \times G$, the mutex check (Definition~\ref{defn:mutex_goal_pair}) involves verifying that each goal is individually reachable and that no reachable state contains both. Individual reachability is in PSPACE by the same nondeterministic argument as for $\mathcal{I}_{\text{UR}}$. The non-coexistence check is in co-PSPACE $=$ PSPACE. Since PSPACE is closed under Boolean combination and there are at most $|G|^2$ pairs, the overall problem is in PSPACE.

    \emph{Hardness.} We reduce from \textsc{PlanSat}. Given $\Pi = \langle P, O, I, G \rangle$, construct $\Pi' = \langle P', O', I', G' \rangle$ where $P' = P \cup \{g_1, g_2, \textit{flag}\}$ for three fresh propositions, $I' = I \cup \{g_1\}$, $G' = \{g_1, g_2\}$, and $O' = O \cup \{o_f, o_a, o_m\}$ with:
    \begin{align*}
        o_f & = \langle \emptyset,\; \{\textit{flag}\},\; \{g_1\} \rangle \\
        o_a & = \langle \{\textit{flag}\},\; \{g_2\},\; \emptyset \rangle \\
        o_m & = \langle G,\; \{g_2\},\; \emptyset \rangle
    \end{align*}
    The original operators in $O$ affect only propositions in $P$ and never modify $g_1$, $g_2$, or $\textit{flag}$. Goal $g_1$ is achievable since it is in $I'$. Goal $g_2$ is always achievable via $o_f$ followed by $o_a$.

    If $\Pi$ is solvable: a plan for $\Pi$ reaches a state satisfying $G$, from which $o_m$ is applicable, producing $g_2$ without deleting $g_1$. The resulting state contains both $g_1$ and $g_2$, so the goals are not mutex, and $\mathcal{I}^{\text{scope}}_{\text{MX}}(\Pi') = 0$.

    If $\Pi$ is unsolvable: $o_m$ is never applicable because its precondition $G$ is never satisfied. The only path to $g_2$ is $o_f \to o_a$, but $o_f$ deletes $g_1$ and no operator in $O'$ re-adds it. No reachable state contains both $g_1$ and $g_2$, so the goals are mutex, and $\mathcal{I}^{\text{scope}}_{\text{MX}}(\Pi') = 2$.

    Therefore $\text{UPPER}_{\mathcal{I}_{\text{MX}}}$ with $k = 0$ decides \textsc{PlanSat}, establishing PSPACE-hardness.
\end{proof}

\begin{cor}
    \label{cor:mx_all}
    $\text{LOWER}_{\mathcal{I}_{\text{MX}}}$ and $\text{EXACT}_{\mathcal{I}_{\text{MX}}}$ are PSPACE-complete. $\text{VALUE}_{\mathcal{I}_{\text{MX}}}$ is in FPSPACE.
\end{cor}
\begin{proof}
    By the same arguments as Corollary~\ref{cor:ur_all}, applied to the reduction from Theorem~\ref{thm:mx_complexity}.
\end{proof}

\begin{thm}
    \label{thm:gs_complexity}
    $\text{UPPER}_{\mathcal{I}_{\text{GS}}}$ is PSPACE-complete.
\end{thm}
\begin{proof}
    \emph{Membership.} The sequencing conflict check (Definition~\ref{defn:sequencing_conflict}) extends the mutex check with an additional condition: for all $g_1$-states $s$, $g_2 \notin \text{REACH}(s)$. This local reachability query is itself in PSPACE by the same nondeterministic search argument. Universally quantifying over $g_1$-states places the check in co-PSPACE $=$ PSPACE. Since all sub-checks are in PSPACE and their combination involves polynomial overhead, the overall problem remains in PSPACE.

    \emph{Hardness.} The reduction from Theorem~\ref{thm:mx_complexity} also establishes hardness for $\mathcal{I}_{\text{GS}}$. In the constructed instance $\Pi'$, when $\Pi$ is unsolvable, no operator in $O'$ adds $g_1$: the only source of $g_1$ is the initial state, and $o_f$ deletes it irreversibly. Thus from any $g_2$-state, which can only be reached via $o_f \to o_a$, $g_1$ is unreachable, making $(g_2, g_1)$ a sequencing conflict. Conversely, when $\Pi$ is solvable, $o_m$ produces a state where both goals co-occur, so no sequencing conflict exists. Therefore $\text{UPPER}_{\mathcal{I}_{\text{GS}}}$ with $k = 0$ decides \textsc{PlanSat}.
\end{proof}

\begin{cor}
    \label{cor:gs_all}
    $\text{LOWER}_{\mathcal{I}_{\text{GS}}}$ and $\text{EXACT}_{\mathcal{I}_{\text{GS}}}$ are PSPACE-complete. $\text{VALUE}_{\mathcal{I}_{\text{GS}}}$ is in FPSPACE.
\end{cor}
\begin{proof}
    By the same arguments as Corollary~\ref{cor:ur_all}, applied to the reduction from Theorem~\ref{thm:gs_complexity}.
\end{proof}

Table~\ref{tab:decision_complexity} summarizes these results.

\begin{table}[ht]
    \centering
    \caption{Complexity of decision and function problems for the proposed planning inconsistency measures. Results hold for both scope and structure variants.}
    \label{tab:decision_complexity}
    \begin{tabular}{l|l|l|l}
        \hline
        Problem        & $\mathcal{I}_{\text{UR}}$ & $\mathcal{I}_{\text{MX}}$ & $\mathcal{I}_{\text{GS}}$ \\
        \hline
        $\text{UPPER}$ & PSPACE-c.                 & PSPACE-c.                 & PSPACE-c.                 \\
        $\text{LOWER}$ & PSPACE-c.                 & PSPACE-c.                 & PSPACE-c.                 \\
        $\text{EXACT}$ & PSPACE-c.                 & PSPACE-c.                 & PSPACE-c.                 \\
        $\text{VALUE}$ & in FPSPACE                & in FPSPACE                & in FPSPACE                \\
        \hline
    \end{tabular}
\end{table}

The uniform PSPACE tier observed above contrasts with the richer complexity landscape for inconsistency measures in other formalisms. Thimm and Wallner~\cite{Thimm2016a} show that propositional inconsistency measures span three distinct tiers, ranging from NP-complete through $\Sigma^p_2$- or $\Pi^p_2$-complete to $\#\cdot\text{coNP}$-complete for counting variants. This stratification arises because the base satisfiability problem is NP-complete, and different measure definitions add varying computational overhead above that base. Corea et al.~\cite{Corea2022} analyze measures for declarative process specifications over $\text{LTL}_{\text{ff}}$, where satisfiability is also NP-complete. Their time-sensitive inconsistency measures consequently land at the first level of the polynomial hierarchy, yielding NP-complete, coNP-complete, $D^p$, and $\text{FP}^{\text{NP}[\log n]}$ problems.

For classical planning, \textsc{PlanSat} is PSPACE-complete, and all three proposed measures reduce to polynomial compositions of reachability queries. Since PSPACE $=$ co-PSPACE, the distinctions between $\text{UPPER}$ and $\text{LOWER}$ and the $D^p$ class for $\text{EXACT}$ all collapse. The polynomial overhead from different measure definitions cannot push the complexity above PSPACE, because PSPACE is closed under polynomial time reductions. The result is a flat complexity landscape: all decision problems for all three measures are PSPACE-complete. This uniformity reflects a general principle: the complexity of inconsistency measures inherits from the base reasoning problem of the target formalism, and when that base problem is already PSPACE-complete, the polynomial differences between measure definitions are subsumed.


The preceding results characterize the inherent complexity of the decision problems. The following propositions complement them with explicit time and space bounds for the enumeration-based implementation.

\begin{prop}[Reachability Complexity]
    \label{prop:reach_complexity}
    Computing the reachable state set $S_{\text{reach}}$ for a planning problem $\langle P, O, I, G \rangle$ requires $O(2^{|P|} \cdot |O| \cdot |P|)$ time and $O(2^{|P|} \cdot |P|)$ space.
\end{prop}
\begin{proof}
    The reachable state set has at most $2^{|P|}$ elements. Each iteration of the fixpoint computation applies up to $|O|$ operators, where each operator application requires $O(|P|)$ for precondition checking and effect computation. Each state requires $O(|P|)$ space.
\end{proof}

\begin{prop}[Unreachability Profile Complexity]
    \label{prop:ur_complexity}
    Given $S_{\text{reach}}$, computing $\mathcal{I}^{\text{scope}}_{\text{UR}}$ and $\mathcal{I}^{\text{struct}}_{\text{UR}}$ requires $O(|S_{\text{reach}}| \cdot |P|)$ time and $O(|P|)$ additional space.
\end{prop}
\begin{proof}
    For each proposition $p \in P$, check whether $p$ appears in any state $s \in S_{\text{reach}}$. This requires one pass through all states.
\end{proof}

\begin{prop}[Goal Mutex Profile Complexity]
    \label{prop:mx_complexity}
    Given $S_{\text{reach}}$, computing $\mathcal{I}^{\text{scope}}_{\text{MX}}$ and $\mathcal{I}^{\text{struct}}_{\text{MX}}$ requires $O(|S_{\text{reach}}| \cdot |G|^2)$ time and $O(|G|^2)$ additional space.
\end{prop}
\begin{proof}
    For each goal pair $(g_1, g_2)$, verify achievability of each goal ($O(|S_{\text{reach}}| \cdot |G|)$) and check no state contains both ($O(|S_{\text{reach}}| \cdot |G|^2)$). Results stored in a $|G| \times |G|$ matrix.
\end{proof}

\begin{prop}[Goal Sequencing Conflict Profile Complexity]
    \label{prop:gs_complexity}
    Given $S_{\text{reach}}$, computing $\mathcal{I}^{\text{scope}}_{\text{GS}}$ and $\mathcal{I}^{\text{struct}}_{\text{GS}}$ requires $O(|G|^2 \cdot |S_{\text{reach}}|^2 \cdot |O|)$ time and $O(|S_{\text{reach}}|)$ additional space.
\end{prop}
\begin{proof}
    Since $\text{REACH}(s) \subseteq S_{\text{reach}}$ for any $s \in S_{\text{reach}}$, local reachability is computed as graph traversal within the pre-computed state space rather than full reachability. For each goal pair $(g_1, g_2)$, check all $g_1$-states; each requires traversing at most $|S_{\text{reach}}|$ states with $|O|$ transitions each. There are $O(|G|^2)$ pairs and up to $|S_{\text{reach}}|$ starting states per pair.
\end{proof}

Table~\ref{tab:complexity} summarizes these results.

\begin{table}[ht]
    \centering
    \caption{Computational complexity of proposed measures}
    \label{tab:complexity}
    \begin{tabular}{l|l|l}
        \hline
        Measure                                              & Time Complexity                                 & Space Complexity        \\
        \hline
        $S_{\text{reach}}$                                   & $O(2^{|P|} \cdot |O| \cdot |P|)$                & $O(2^{|P|} \cdot |P|)$  \\
        $\mathcal{I}_{\text{UR}}$ (given $S_{\text{reach}}$) & $O(|S_{\text{reach}}| \cdot |P|)$               & $O(|P|)$                \\
        $\mathcal{I}_{\text{MX}}$ (given $S_{\text{reach}}$) & $O(|S_{\text{reach}}| \cdot |G|^2)$             & $O(|G|^2)$              \\
        $\mathcal{I}_{\text{GS}}$ (given $S_{\text{reach}}$) & $O(|G|^2 \cdot |S_{\text{reach}}|^2 \cdot |O|)$ & $O(|S_{\text{reach}}|)$ \\
        \hline
    \end{tabular}
\end{table}

The exponential dependence on $|P|$ reflects the inherent complexity of planning. However, real planning problems often have much smaller reachable state spaces than the theoretical maximum. For example, the Locked Door scenario has $|P| = 4$ but $|S_{\text{reach}}| = 1$. Additionally, all three measures share the $S_{\text{reach}}$ computation, amortizing this cost when computing multiple profiles. The Goal Sequencing Conflict Profile has higher complexity than the other measures due to requiring multiple local reachability computations. For expensive problems, the Unreachability and Mutex Profiles can be computed independently as partial diagnostics.

The complexity analysis above assumes explicit state space enumeration. \ac{ASP}-based implementations may exhibit different practical performance characteristics due to several factors. First, modern \ac{ASP} solvers employ conflict-driven learning, unit propagation, and backjumping, reducing inferences to nogood propagation and enabling efficient search space traversal~\cite{Gebser2012}. Second, the declarative nature of \ac{ASP} encodings allows solvers to exploit problem structure automatically.

Consequently, the theoretical worst-case bounds may significantly overestimate practical computation times. The empirical evaluation (Chapter~\ref{sec:evaluation}) will assess actual scalability on benchmark problems, providing complementary evidence to the theoretical complexity analysis. Kuhlmann et al.~\cite{Kuhlmann2025} demonstrate that \ac{ASP}-based implementations significantly outperform both \ac{SAT}-based and naive baseline approaches for computing propositional inconsistency measures, suggesting that similar advantages may apply to the planning-based measures proposed here.
